---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ineqTrees

<!-- badges: start -->

 [![R-CMD-check](https://github.com/m-mburu/ineqTrees/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/m-mburu/ineqTrees/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->



`ineqTrees` provides tools for studying socioeconomic inequality in health
outcomes with tree-based models. The package includes weighted rank and
concentration-index utilities, inequality-aware split scoring, and wrappers
around `partykit` for fitting conditional inference trees and forests.

## Installation

You can install the development version of ineqTrees like so:

``` r
remotes::install_github("m-mburu/ineqTrees")
```

## Fitting a tree

The example below fits an inequality-aware conditional inference tree on a
sample from the built-in `kenya` dataset. The response combines the ranking
variable (`wealth`) and the health outcome (`deadu5_num`), while the split
criterion is based on concentration-index reduction.

```{r example, fig.width=10, fig.height=6}
pkgload::load_all(quiet = TRUE)
load("data/kenya.rda")

set.seed(1)

toy_kenya <- kenya[sample(nrow(kenya), 500), ]
toy_kenya$reg <- droplevels(toy_kenya$reg)

fit_tree <- ctree_ci(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg,
  data = toy_kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  control = partykit::ctree_control(mincriterion = 0.95, maxdepth = 4)
)

plot(
  fit_tree,
  gp = grid::gpar(fontsize = 6.5),
  data = toy_kenya,
  terminal_stats = list(
    death_rate = function(df) mean(df$deadu5_num),
    mean_wealth = function(df) mean(df$wealth),
    concentration_index = function(df) {
      ci_factory("CI")(cbind(df$wealth, df$deadu5_num), rep(1, nrow(df)))
    }
  ),
  stat_labels = c(
    death_rate = "Death rate",
    mean_wealth = "Mean wealth",
    concentration_index = "CI"
  ),
  stat_formatters = list(
    death_rate = function(x) sprintf("%.3f", x),
    mean_wealth = function(x) sprintf("%.2f", x),
    concentration_index = function(x) sprintf("%.3f", x)
  ),
  tp_args = list(
    width_lines = 10.5,
    height_lines = 4.2
  ),
  tnex = 0.85
)
```

## Fit forests

```{r helpers}
fit_forest <- cf_ci(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg,
  data = toy_kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  ntree = 10L,
  mtry = 1L,
  control = partykit::ctree_control(mincriterion = 0.95, maxdepth = 4)
)
```


### Fit surrogate trees

```{r surrogate, fig.width=10, fig.height=6}
preds <- stats::predict(
  fit_forest,
  newdata = toy_kenya,
  OOB = FALSE,
  FUN = function(y, w) stats::weighted.mean(y[, "deadu5_num"], w)
)

surrogate_data <- transform(toy_kenya, forest_risk = preds)

surrogate_fit <- ctree_ci(
  cbind(wealth, forest_risk) ~ rural + ed + unskilled,
  data = surrogate_data,
  rank_name = "wealth",
  outcome_name = "forest_risk",
  control = partykit::ctree_control(mincriterion = 0.95, maxdepth = 4)
)

summary(preds)
plot(
  surrogate_fit,
  gp = grid::gpar(fontsize = 6.5),
  data = surrogate_data,
  terminal_stats = list(
    predicted_death_rate = function(df) mean(df$forest_risk),
    mean_wealth = function(df) mean(df$wealth),
    concentration_index = function(df) {
      ci_factory("CI")(cbind(df$wealth, df$forest_risk), rep(1, nrow(df)))
    }
  ),
  stat_labels = c(
    predicted_death_rate = "death rate",
    mean_wealth = "Mean wealth",
    concentration_index = "CI"
  ),
  stat_formatters = list(
    predicted_death_rate = function(x) sprintf("%.3f", x),
    mean_wealth = function(x) sprintf("%.2f", x),
    concentration_index = function(x) sprintf("%.3f", x)
  ),
  tp_args = list(
    width_lines = 10.5,
    height_lines = 4.2
  ),
  tnex = 0.85
)
```
