---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 1000
)
```

# ineqTrees

<!-- badges: start -->

 [![R-CMD-check](https://github.com/m-mburu/ineqTrees/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/m-mburu/ineqTrees/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->



`ineqTrees` provides tools for studying socioeconomic inequality in health
outcomes with tree-based models. The package includes weighted rank and
concentration-index utilities, inequality-aware split scoring, and wrappers
around `partykit` for fitting conditional inference trees and forests.

## Installation

You can install the development version of ineqTrees like so:

``` r
remotes::install_github("m-mburu/ineqTrees")
```

## Fitting a tree

The example below fits an inequality-aware conditional inference tree on a
sample from the built-in `kenya` dataset. The response combines the ranking
variable (`wealth`) and the health outcome (`deadu5_num`), while the split
criterion is based on concentration-index reduction.

```{r example, fig.width=15, fig.height=9}
library(ineqTrees)
load("data/kenya.rda")

set.seed(1)

fit_tree <- ctree_ci(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  control = partykit::ctree_control(mincriterion = 0.95, maxdepth = 4)
)

plot(
  fit_tree,
  gp = grid::gpar(fontsize = 6.5),
  data = kenya,
  var_labels = c(
    rural = "Residence",
    reg = "Province",
    ed = "Mother education",
    unskilled = "Mother occupation"
  ),
  terminal_stats = list(
    n = nrow,
    death_rate = function(df) mean(df$deadu5_num),
    mean_wealth = function(df) mean(df$wealth),
    ci = function(df) {
      ci_factory("CI")(cbind(df$wealth, df$deadu5_num), rep(1, nrow(df)))
    }
  ),
  stat_labels = list(
    n = "n",
    death_rate = "% death",
    mean_wealth = expression(mu ~ wealth),
    ci = "CI"
  ),
  stat_formatters = list(
    n = function(x) format(x, big.mark = ",", scientific = FALSE),
    death_rate = function(x) sprintf("%.2f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x),
    ci = function(x) sprintf("%.3f", x)
  ),
  terminal_fill = "#d9d9d9",
  tp_args = list(
    width_lines = 11,
    height_lines = 5.2
  ),
  tnex = 0.85
)
```

## Fit forests

```{r randomForest}
fit_forest <- cf_ci(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  ntree = 10L,
  mtry = 1L,
  control = partykit::ctree_control(mincriterion = 0.95, maxdepth = 4)
)
```


### Fit surrogate trees

```{r surrogate, fig.width=15, fig.height=9}
preds <- stats::predict(
  fit_forest,
  newdata = kenya,
  OOB = FALSE,
  FUN = function(y, w) stats::weighted.mean(y[, "deadu5_num"], w)
)

surrogate_data <- transform(kenya, forest_risk = preds)

surrogate_fit <- ctree_ci(
  cbind(wealth, forest_risk) ~ rural + ed + reg + unskilled,
  data = surrogate_data,
  rank_name = "wealth",
  outcome_name = "forest_risk",
  control = partykit::ctree_control(mincriterion = 0.95, maxdepth = 4)
)

summary(preds)
plot(
  surrogate_fit,
  gp = grid::gpar(fontsize = 6.5),
  data = surrogate_data,
  var_labels = c(
    rural = "Residence",
    reg = "Province",
    ed = "Mother education",
    unskilled = "Mother occupation"
  ),
  terminal_stats = list(
    n = nrow,
    predicted_death_rate = function(df) mean(df$forest_risk),
    mean_wealth = function(df) mean(df$wealth),
    ci = function(df) {
      ci_factory("CI")(cbind(df$wealth, df$forest_risk), rep(1, nrow(df)))
    }
  ),
  stat_labels = list(
    n = "n",
    predicted_death_rate = "% death",
    mean_wealth = expression(mu ~ wealth),
    ci = "CI"
  ),
  stat_formatters = list(
    n = function(x) format(x, big.mark = ",", scientific = FALSE),
    predicted_death_rate = function(x) sprintf("%.2f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x),
    ci = function(x) sprintf("%.3f", x)
  ),
  terminal_fill = "#d9d9d9",
  tp_args = list(
    width_lines = 11,
    height_lines = 5.2
  ),
  tnex = 0.85
)
```

## SHAP-based decomposition

- Approximate SHAP values for the fitted forest with `fastshap::explain()`, using a prediction wrapper that returns the predicted outcome for each observation. 
- Decompose the concentration index of those predicted risks with `shap_conc_decomp()`.

```{r shap-decomposition, fig.width=8, fig.height=5}
predict_cf_ci_wrapper <- function(object, newdata) {
  stats::predict(
    object,
    newdata = newdata,
    OOB = FALSE,
    FUN = function(y, w) stats::weighted.mean(y[, "deadu5_num"], w)
  )
}

set.seed(20260328)
shap_eval_n <- min(400L, nrow(kenya))
shap_rows <- sort(sample.int(nrow(kenya), shap_eval_n))
forest_X <- kenya[c("rural", "ed", "reg", "unskilled")]
shap_X_eval <- forest_X[shap_rows, , drop = FALSE]
forest_pred_eval <- predict_cf_ci_wrapper(fit_forest, shap_X_eval)
wealth_eval <- kenya$wealth[shap_rows]

forest_shap <- fastshap::explain(
  object = fit_forest,
  X = forest_X,
  pred_wrapper = predict_cf_ci_wrapper,
  newdata = shap_X_eval,
  nsim = 64,
  adjust = TRUE
)

decomp <- shap_conc_decomp(
  shap = forest_shap,
  rank = wealth_eval,
  prediction = forest_pred_eval
)

decomp$diagnostics
decomp$contributions

ggplot2::ggplot(
  as.data.frame(decomp$contributions),
  ggplot2::aes(
    x = stats::reorder(feature, pct_contribution),
    y = pct_contribution,
    fill = pct_contribution > 0
  )
) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(
    values = c("#2166ac", "#b2182b"),
    guide = "none"
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Percentage contribution",
    title = "SHAP-based concentration-index decomposition"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())
```
