---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 600
)
```

# ineqTrees

<!-- badges: start -->

 [![R-CMD-check](https://github.com/m-mburu/ineqTrees/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/m-mburu/ineqTrees/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->



`ineqTrees` provides tools for studying socioeconomic inequality in health
outcomes with tree-based models. The package includes weighted rank and
concentration-index utilities, inequality-aware split scoring, and wrappers
for fitting greedy concentration-index trees and forests.

## Installation

You can install the development version of ineqTrees like so:

``` r
remotes::install_github("m-mburu/ineqTrees")
```

## Fitting a tree

The example below fits an inequality-aware greedy tree on a
sample from the built-in `kenya` dataset. The response combines the ranking
variable (`wealth`) and the health outcome (`deadu5_num`), while the split
criterion is based on concentration-index reduction.


```{r readme-helpers, include=FALSE}
readme_predictors <- c("rural", "ed", "reg", "unskilled")

readme_tree_plot <- function(fit, data, outcome_name, outcome_label = "% death") {
  metric_name <- "outcome_rate"
  outcome_stat <- function(df) mean(df[[outcome_name]], na.rm = TRUE)

  plot(
    fit,
    gp = grid::gpar(fontsize = 6.5),
    data = data,
    var_labels = c(
      rural = "Residence",
      reg = "Province",
      ed = "Mother education",
      unskilled = "Mother occupation"
    ),
    terminal_stats = c(
      list(n = nrow),
      stats::setNames(list(outcome_stat), metric_name),
      list(
        mean_wealth = function(df) mean(df$wealth, na.rm = TRUE),
        ci = function(df) {
          ci_factory("CI")(
            cbind(df$wealth, df[[outcome_name]]),
            rep(1, nrow(df))
          )
        }
      )
    ),
    stat_labels = c(
      list(n = "n"),
      stats::setNames(list(outcome_label), metric_name),
      list(
        mean_wealth = expression(mu ~ wealth),
        ci = "CI"
      )
    ),
    stat_formatters = c(
      list(
        n = function(x) format(x, big.mark = ",", scientific = FALSE)
      ),
      stats::setNames(
        list(function(x) sprintf("%.2f%%", 100 * x)),
        metric_name
      ),
      list(
        mean_wealth = function(x) sprintf("%.2f", x),
        ci = function(x) sprintf("%.3f", x)
      )
    ),
    terminal_fill = "#d9d9d9",
    tp_args = list(
      width_lines = 11,
      height_lines = 5.2
    ),
    tnex = 0.85
  )
}

readme_forest_predict <- function(object, newdata) {
  as.numeric(
    stats::predict(
      object,
      newdata = as.data.frame(newdata),
      OOB = FALSE
    )
  )
}

readme_tuning_table <- function(x, columns, labels) {
  out <- as.data.frame(x)
  missing_columns <- setdiff(columns, names(out))
  for (column in missing_columns) {
    out[[column]] <- NA
  }
  out <- out[, columns, drop = FALSE]
  names(out) <- labels
  out
}

```

### load data and set seed for reproducibility

```{r readme-data}
if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  suppressMessages(pkgload::load_all(export_all = FALSE))
} else {
  library(ineqTrees)
}
library(data.table)
load("data/kenya.rda")

set.seed(1)
```

### Fit  tree

- This is a concentration-index tree, so the response is a two-column matrix of the ranking variable and the outcome. The `rank_name` and `outcome_name` arguments specify which columns of the data to use for those roles. The split criterion is based on concentration-index reduction, so the `control` argument specifies greedy controls rather than conditional-inference test controls.

```{r fit-tree}
fit_tree <- ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  control = ci_tree_control(maxdepth = 4L)
)
```

```{r tree-summary}
#fit_tree
ci_tree_terminal_summary(fit_tree)
```

```{r readme-tree-plot, fig.width=15, fig.height=9, fig.alt="A concentration-index tree fitted to under-five mortality in the Kenya example data."}
readme_tree_plot(fit_tree, kenya, "deadu5_num")
```


## Fitting a forest

The forest interface uses the same response specification, but averages
predictions across many greedy concentration-index trees. The tuned workflow
later in the README uses the same model family with cross-validation.

```{r fit-forest}
fit_forest <- ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  ntree = 10L,
  mtry = 1L,
  control = ci_tree_control(maxdepth = 5L)
)
fit_forest
```

```{r}
ci_forest_summary(fit_forest)
```

### Parallel execution patterns

`ineqTrees` uses the `future` ecosystem for optional parallel work and never
sets the future plan inside model functions. Choose one layer at a time: direct
forest fitting, custom tuning tasks, or tidymodels resamples.

```r
old_plan <- future::plan()
future::plan(future::multisession, workers = 2)
on.exit(future::plan(old_plan), add = TRUE)

fit_forest <- ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  ntree = 50L,
  parallel = TRUE
)
```

```r
old_plan <- future::plan()
future::plan(future::multisession, workers = 2)
on.exit(future::plan(old_plan), add = TRUE)

forest_tune_grid <- ci_tree_control_grid(
  minsplit = 100L,
  minbucket = c(50L, 100L),
  maxdepth = 3:4,
  mtry = c(1L, 2L),
  ntree = c(10L, 50L)
)

forest_tuning <- tune_ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  control_grid = forest_tune_grid,
  v = 5L,
  parallel_over = "tuning"
)
```

```r
old_plan <- future::plan()
future::plan(future::multisession, workers = 2)
on.exit(future::plan(old_plan), add = TRUE)

forest_tuned <- tune::tune_grid(
  forest_wf,
  resamples = tree_folds,
  grid = forest_grid,
  metrics = yardstick::metric_set(ci_gain_tuning, yardstick::rmse),
  control = tune::control_grid(parallel_over = "resamples")
)
```

### Fit surrogarete tree to forest predictions

- The surrogate is a greedy concentration-index tree that approximates the predictions of the fitted forest. This is useful for interpreting the forest , since the surrogate can be refit on held-out data and scored with CI gain.

```{r surrogate-tree}
setDT(kenya)
kenya[, forest_risk := readme_forest_predict(
   fit_forest, .SD
), .SDcols = readme_predictors]

surrogate_tree <- ci_tree(
  formula = cbind(wealth, forest_risk) ~ rural + ed + reg + unskilled,
  data = kenya,
  rank_name = "wealth",
  outcome_name = "forest_risk",
  control = ci_tree_control(maxdepth = 4L)
)
#surrogate_tree
```

## plot

```{r surrogate-tree-plot, fig.width=15, fig.height=9, fig.alt="A surrogate concentration-index tree approximating fitted forest predictions."}
readme_tree_plot(
  surrogate_tree,
  kenya,
  outcome_name = "forest_risk",
  outcome_label = "Predicted risk"
)
```



## SHAP-based decomposition

- Approximate SHAP values for the fitted forest with `fastshap::explain()`, using a prediction wrapper that returns the predicted outcome for each observation. 
- Decompose the concentration index of those predicted risks with `shap_conc_decomp()`.

The model does not have to be an `ineqTrees` forest. `shap_conc_decomp()` only
needs three things: SHAP values, the socioeconomic ranking variable, and the
model's predicted risk. That means the same step also works after fitting an
ordinary prediction model, for example a `ranger` random forest tuned with
tidymodels for risk of under-five death.

```{r ranger-tidymodels-shap-sketch}
library(tidymodels)
library(ranger)

ranger_data <- as.data.frame(
  kenya[, c("wealth", "deadu5_num", readme_predictors), with = FALSE]
)
ranger_data$deadu5 <- factor(
  ifelse(ranger_data$deadu5_num == 1, "died", "alive"),
  levels = c("died", "alive")
)

ranger_risk_predict <- function(object, newdata) {
  prob <- predict(object, new_data = as.data.frame(newdata), type = "prob")
  as.numeric(prob$.pred_died)
}

ranger_folds <- vfold_cv(ranger_data, v = 5, strata = deadu5)

ranger_spec <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", probability = TRUE) %>%
  set_mode("classification")

ranger_wflow <- workflow() %>%
  add_model(ranger_spec) %>%
  add_formula(deadu5 ~ rural + ed + reg + unskilled)

ranger_grid <- grid_regular(
  mtry(range = c(1L, length(readme_predictors))),
  min_n(range = c(5L, 40L)),
  levels = 3
)

ranger_tuning <- tune_grid(
  ranger_wflow,
  resamples = ranger_folds,
  grid = ranger_grid,
  metrics = metric_set(roc_auc, accuracy)
)

best_ranger <- finalize_workflow(
  ranger_wflow,
  select_best(ranger_tuning, metric = "roc_auc")
) %>%
  fit(data = ranger_data)

ranger_X <- ranger_data[, readme_predictors, drop = FALSE]
ranger_eval_n <- min(400L, nrow(ranger_data))
ranger_rows <- sort(sample.int(nrow(ranger_data), ranger_eval_n))
ranger_X_eval <- ranger_X[ranger_rows, , drop = FALSE]
ranger_pred_eval <- ranger_risk_predict(best_ranger, ranger_X_eval)

ranger_shap <- fastshap::explain(
  object = best_ranger,
  X = ranger_X,
  pred_wrapper = ranger_risk_predict,
  newdata = ranger_X_eval,
  nsim = 64,
  adjust = TRUE
)

ranger_decomp <- shap_conc_decomp(
  shap = ranger_shap,
  rank = ranger_data$wealth[ranger_rows],
  prediction = ranger_pred_eval
)
```

```{r ranger-shap-decomposition-data}
ranger_shap_diagnostics <- as.data.frame(ranger_decomp$diagnostics)
ranger_shap_contrib_table <- as.data.frame(ranger_decomp$contributions)
ranger_shap_contrib_table <- ranger_shap_contrib_table[
  order(-ranger_shap_contrib_table$abs_contribution),
  ,
  drop = FALSE
]
```

```{r ranger-shap-diagnostics}
knitr::kable(
  ranger_shap_diagnostics,
  digits = 3,
  caption = "Ranger SHAP decomposition diagnostics"
)
```

```{r ranger-shap-contributions-table}
knitr::kable(
  ranger_shap_contrib_table,
  digits = 3,
  caption = "Ranger SHAP-based concentration-index contributions"
)
```

```{r ranger-shap-decomposition-plot, fig.width=8, fig.height=5, fig.alt="A horizontal bar chart of ranger SHAP-based percentage contributions to the concentration index."}
library(ggplot2)
ggplot(
  ranger_shap_contrib_table,
  aes(
    x = stats::reorder(feature, pct_contribution),
    y = pct_contribution,
    fill = pct_contribution > 0
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(
    values = c("#2166ac", "#b2182b"),
    guide = "none"
  ) +
  labs(
    x = NULL,
    y = "Percentage contribution",
    title = "Ranger SHAP-based concentration-index decomposition"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

```{r shap-compute}
set.seed(20260328)
shap_eval_n <- min(400L, nrow(kenya))
shap_rows <- sort(sample.int(nrow(kenya), shap_eval_n))
forest_X <- kenya[, ..readme_predictors]
shap_X_eval <- forest_X[shap_rows, , drop = FALSE]
forest_pred_eval <- readme_forest_predict(fit_forest, shap_X_eval)
wealth_eval <- kenya$wealth[shap_rows]

forest_shap <- fastshap::explain(
  object = fit_forest,
  X = forest_X,
  pred_wrapper = readme_forest_predict,
  newdata = shap_X_eval,
  nsim = 64,
  adjust = TRUE
)

decomp <- shap_conc_decomp(
  shap = forest_shap,
  rank = wealth_eval,
  prediction = forest_pred_eval
)

shap_diagnostics <- as.data.frame(decomp$diagnostics)
shap_contrib_table <- as.data.frame(decomp$contributions)
shap_contrib_table <- shap_contrib_table[
  order(-shap_contrib_table$abs_contribution),
  ,
  drop = FALSE
]
```

```{r shap-diagnostics}
knitr::kable(
  shap_diagnostics,
  digits = 3,
  caption = "SHAP decomposition diagnostics"
)
```

```{r shap-contributions-table}
knitr::kable(
  shap_contrib_table,
  digits = 3,
  caption = "SHAP-based concentration-index contributions"
)
```

```{r shap-decomposition-plot, fig.width=8, fig.height=5, fig.alt="A horizontal bar chart of SHAP-based percentage contributions to the concentration index."}
library(ggplot2)
ggplot(
  shap_contrib_table,
  aes(
    x = stats::reorder(feature, pct_contribution),
    y = pct_contribution,
    fill = pct_contribution > 0
  )
) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(
    values = c("#2166ac", "#b2182b"),
    guide = "none"
  ) +
  labs(
    x = NULL,
    y = "Percentage contribution",
    title = "SHAP-based concentration-index decomposition"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```


```{r tuning-data}
set.seed(20260507)
tuning_n <- min(800L, nrow(kenya))
tuning_rows <- sort(sample.int(nrow(kenya), tuning_n))
tuning_data <- kenya[tuning_rows, , drop = FALSE]
```

## Tune tree hyperparameters

The current model-selection workflow uses `ci_tree_control_grid()` to define
candidate greedy controls and `tune_ci_tree()` to score them with
cross-validation. The concentration-index variant is tuned alongside the tree
controls by passing several values to `type`.

```{r tree-tuning-grid}
tree_tune_grid <- ci_tree_control_grid(
  minsplit = c(100L),
  minbucket = c(50L, 100L),
  maxdepth = c(3L:6L),
  minprob = c(0.01, 0.1)
)
```

```{r tree-tuning-fit}
tree_tuning <- tune_ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = tuning_data,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  type = c("CI", "CIg", "CIc", "L"),
  control_grid = tree_tune_grid,
  v = 5L,
  strata = "deadu5_num",
  seed = 20260507,
  metric = "validation_gain",
  refit = TRUE
)
```

```{r tree-tuning-table}
tree_tuning_table <- readme_tuning_table(
  tree_tuning$summary,
  columns = c(
    "type",
    "minsplit",
    "minbucket",
    "maxdepth",
    "mean_score",
    "sd_score",
    "mean_terminal_nodes"
  ),
  labels = c(
    "type",
    "minsplit",
    "minbucket",
    "maxdepth",
    "mean_validation_gain",
    "sd_validation_gain",
    "mean_terminal_nodes"
  )
)

knitr::kable(
  tree_tuning_table[1:5,],
  digits = 3,
  caption = "Cross-validated greedy tree tuning results"
)
```

```{r tree-tuning-plot, fig.width=15, fig.height=9, fig.alt="The best concentration-index tree selected by cross-validated tuning."}
readme_tree_plot(
  fit = tree_tuning$best_fit,
  data = tuning_data,
  outcome_name = "deadu5_num"
)
```

## Tune forest hyperparameters

For forests, `tune_ci_forest()` uses the same greedy controls and adds `ntree` when
that column is present in the tuning grid. Each candidate forest is summarized
by a surrogate greedy CI tree, and the grid is ranked by held-out CI validation
gain from that surrogate.

```{r forest-tuning-grid}
forest_tune_grid <- ci_tree_control_grid(
  minsplit = c(100L),
  minbucket = c(50L, 100L),
  maxdepth = c(3L:6L),
  mtry = c(1L, 2L),
  ntree = c(10L, 50L),
  minprob = c(0.01, 0.1)
)
```

```{r forest-tuning-fit}
forest_tuning <- tune_ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = tuning_data,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  type = c("CI", "CIg", "CIc", "L"),
  control_grid = forest_tune_grid,
  v = 5L,
  strata = "deadu5_num",
  seed = 20260508,
  prediction_name = "forest_risk",
  parallel_over = "tuning",
  refit = TRUE
)

```

```{r forest-tuning-table}
forest_tuning_table <- readme_tuning_table(
  forest_tuning$summary,
  columns = c(
    "type",
    "ntree",
    "mtry",
    "maxdepth",
    "mean_score",
    "sd_score",
    "mean_terminal_nodes"
  ),
  labels = c(
    "type",
    "ntree",
    "mtry",
    "maxdepth",
    "mean_validation_gain",
    "sd_validation_gain",
    "mean_terminal_nodes"
  )
)

knitr::kable(
  forest_tuning_table[1:5, ],
  digits = 3,
  caption = "Cross-validated greedy forest tuning results ranked by validation gain"
)
```

```{r forest-surrogate-data}
best_tuned_forest <- forest_tuning$best_fit
forest_surrogate_data <- forest_tuning$best_surrogate_data
```

```{r forest-surrogate-fit}
forest_surrogate_fit <- forest_tuning$best_surrogate
```

```{r forest-surrogate-plot, fig.width=15, fig.height=9, fig.alt="A surrogate tree summarizing the best tuned concentration-index forest."}
readme_tree_plot(
  fit = forest_surrogate_fit,
  data = forest_surrogate_data,
  outcome_name = "forest_risk",
  outcome_label = "Predicted risk"
)
```
