---
title: "CI split search with Rcpp"
output: github_document
---

This development file documents the Rcpp-backed split-search fast path. It is
kept separate from `dev/ci_split_search.Rmd` so the original R implementation
remains available as the reference implementation.

Package files owned by this development note:

- `R/ci_split_search_cpp.R`
- `src/ci_split_search_cpp.cpp`
- `R/RcppExports.R`
- `src/RcppExports.cpp`
- `tests/testthat/test-ci_split_search_cpp.R`

The exported R-facing functions mirror the existing split-search helpers and
add the `_cpp` suffix:

- `best_numeric_split_cpp()`
- `best_factor_split_cpp()`
- `best_split_for_one_variable_cpp()`
- `best_global_ci_split_cpp()`

The global helper also accepts `vars`, an optional integer vector of predictor
ids to search. When `vars` is omitted, all columns in `x` are searched.

The roxygen source for the generated wrapper should document this argument as:

```r
#' @param vars Optional integer variable ids to search. Defaults to all columns
#'   of `x`.
```

The current production tree and forest builders still call the R split-search
helpers. The `_cpp` helpers are a parallel fast path for testing and later
integration.

```{r, function-ci_split_search_cpp_wrappers}
#' Greedy CI split search using Rcpp engines
#'
#' @description
#' Experimental Rcpp-backed versions of the greedy concentration-index split
#' search helpers. These functions keep the same return shape as the current R
#' split helpers, but take `type` instead of a prebuilt `ci_fun` because the
#' concentration-index scoring is evaluated in C++.
#'
#' @param x A numeric predictor vector.
#' @param x_full A factor predictor containing all original levels.
#' @param keep A logical vector selecting the observations currently available
#'   for splitting.
#' @param y,y_full A two-column numeric response matrix.
#' @param wt,wt_full Case weights.
#' @param varid Integer identifier of the splitting variable.
#' @param ctrl A control list created by [ci_tree_control()] or compatible.
#' @param type One of `"CI"`, `"CIg"`, `"CIc"`, or `"L"`.
#' @param return Either `"split"` or `"candidate"`.
#' @param vars Optional integer variable ids to search. Defaults to all columns
#'   of `x`.
#'
#' @return A `partysplit` object, split candidate list, or `NULL`.
#'
#' @useDynLib ineqTrees, .registration = TRUE
#' @importFrom Rcpp sourceCpp
#' @name ci_split_search_cpp
NULL

#' @rdname ci_split_search_cpp
#' @export
best_numeric_split_cpp <- function(
    x,
    y,
    wt,
    varid,
    ctrl,
    type = c("CI", "CIg", "CIc", "L"),
    return = c("split", "candidate")) {
  type <- match.arg(type)
  return <- match.arg(return)
  ctrl <- .ci_tree_normalize_control(ctrl)

  stopifnot(is.matrix(y), ncol(y) == 2L, length(x) == nrow(y),
            length(wt) == nrow(y))

  if (identical(type, "L")) {
    .ci_warn_negative_l_levels(y[, 1])
  }

  out <- ci_best_numeric_split_cpp_engine(
    x = as.numeric(x),
    y = y,
    wt = as.numeric(wt),
    minbucket = ctrl$minbucket,
    minprob = ctrl$minprob,
    type = type
  )

  if (!is.finite(out$gain)) {
    return(NULL)
  }

  sp <- partykit::partysplit(
    varid = as.integer(varid),
    breaks = out$cutpoint,
    index = 1:2
  )

  if (identical(return, "split")) {
    return(sp)
  }

  list(
    gain = out$gain,
    varid = as.integer(varid),
    partysplit = sp,
    split = sp,
    left = as.logical(out$left),
    type = "numeric",
    cutpoint = out$cutpoint
  )
}

#' @rdname ci_split_search_cpp
#' @export
best_factor_split_cpp <- function(
    x_full,
    keep,
    y_full,
    wt_full,
    varid,
    ctrl,
    type = c("CI", "CIg", "CIc", "L"),
    return = c("split", "candidate")) {
  type <- match.arg(type)
  return <- match.arg(return)
  ctrl <- .ci_tree_normalize_control(ctrl)

  if (!is.factor(x_full)) {
    x_full <- factor(x_full)
  }
  stopifnot(is.matrix(y_full), ncol(y_full) == 2L,
            length(x_full) == nrow(y_full), length(wt_full) == nrow(y_full),
            length(keep) == nrow(y_full))

  full_levels <- levels(x_full)
  x <- x_full[keep]
  y <- y_full[keep, , drop = FALSE]
  wt <- wt_full[keep]

  if (identical(type, "L")) {
    .ci_warn_negative_l_levels(y[, 1])
  }

  code <- as.integer(x)
  out <- ci_best_factor_split_cpp_engine(
    code = code,
    y = y,
    wt = as.numeric(wt),
    n_levels = length(full_levels),
    minbucket = ctrl$minbucket,
    minprob = ctrl$minprob,
    type = type
  )

  if (!is.finite(out$gain)) {
    return(NULL)
  }

  left_codes <- as.integer(out$left_codes)
  present_codes <- sort(unique(code[!is.na(code)]))
  right_codes <- setdiff(present_codes, left_codes)

  index <- rep(NA_integer_, length(full_levels))
  index[left_codes] <- 1L
  index[right_codes] <- 2L

  sp <- partykit::partysplit(
    varid = as.integer(varid),
    index = index
  )

  if (identical(return, "split")) {
    return(sp)
  }

  list(
    gain = out$gain,
    varid = as.integer(varid),
    partysplit = sp,
    split = sp,
    left = as.logical(out$left),
    type = "factor",
    left_levels = full_levels[left_codes],
    right_levels = full_levels[right_codes]
  )
}

#' @rdname ci_split_search_cpp
#' @export
best_split_for_one_variable_cpp <- function(
    x,
    y,
    wt,
    varid,
    ctrl,
    type = c("CI", "CIg", "CIc", "L")) {
  type <- match.arg(type)
  stopifnot(length(x) == nrow(y), length(wt) == nrow(y))
  ctrl <- .ci_tree_normalize_control(ctrl)

  keep <- stats::complete.cases(y) & !is.na(x) & !is.na(wt) & wt > 0
  if (sum(wt[keep], na.rm = TRUE) < ctrl$minsplit) {
    return(NULL)
  }

  candidate <- NULL

  if (is.numeric(x) || is.integer(x)) {
    candidate <- best_numeric_split_cpp(
      x = x[keep],
      y = y[keep, , drop = FALSE],
      wt = wt[keep],
      varid = varid,
      ctrl = ctrl,
      type = type,
      return = "candidate"
    )
  } else if (is.factor(x) || is.ordered(x)) {
    candidate <- best_factor_split_cpp(
      x_full = x,
      keep = keep,
      y_full = y,
      wt_full = wt,
      varid = varid,
      ctrl = ctrl,
      type = type,
      return = "candidate"
    )
  }

  if (is.null(candidate)) {
    return(NULL)
  }

  left <- rep(FALSE, length(x))
  left[keep] <- candidate$left
  candidate$left <- left
  candidate
}

#' @rdname ci_split_search_cpp
#' @export
best_global_ci_split_cpp <- function(x,
                                     y,
                                     wt,
                                     ctrl,
                                     type = c("CI", "CIg", "CIc", "L"),
                                     vars = seq_along(x)) {
  type <- match.arg(type)
  stopifnot(is.matrix(y), nrow(y) == length(wt))
  ctrl <- .ci_tree_normalize_control(ctrl)

  vars <- as.integer(vars)
  vars <- vars[vars %in% seq_along(x)]

  if (!length(vars)) {
    return(NULL)
  }

  if (!is.null(ctrl$mtry)) {
    vars <- sample(vars, min(ctrl$mtry, length(vars)))
  }

  parent_impurity <- ci_factory(type)(y, wt)
  best <- NULL

  for (j in vars) {
    candidate <- best_split_for_one_variable_cpp(
      x = x[[j]],
      y = y,
      wt = wt,
      varid = j,
      ctrl = ctrl,
      type = type
    )

    if (!is.null(candidate) &&
        (is.null(best) || candidate$gain > best$gain)) {
      best <- candidate
    }
  }

  if (is.null(best) ||
      !is.finite(best$gain) ||
      best$gain <= ctrl$min_gain) {
    return(NULL)
  }

  best$parent_impurity <- parent_impurity
  best$relative_gain <- .ci_relative_gain(best$gain, parent_impurity)

  if (ctrl$min_relative_gain > 0 &&
      (!is.finite(best$relative_gain) ||
       best$relative_gain <= ctrl$min_relative_gain)) {
    return(NULL)
  }

  best
}
```

```{r cpp-wrapper-example, eval=FALSE}
best_global_ci_split_cpp(
  x = predictors,
  y = y,
  wt = wt,
  ctrl = ci_tree_control(),
  type = "CI"
)
```

## Design notes

The C++ engine computes the same concentration-index gain as the R path:

1. Compute the parent CI once.
2. Search all admissible candidate splits.
3. Compute child-specific weighted fractional ranks inside each candidate child.
4. Score child nodes with the direct population-style weighted covariance,
   without a child-specific finite-sample correction.
5. Return the best gain and child membership.

The C++ functions return compact engine results. The R wrappers convert those
results into the same user-facing objects as the R split search: `partysplit`
objects for `return = "split"` and candidate lists for `return = "candidate"`.

## Verification

The tests compare the C++ path against the current R implementation for:

- numeric split search,
- factor split search with unused levels,
- mixed-predictor global split search,
- tied socioeconomic rank values.

```{r tests-ci_split_search_cpp_tied_ranks}
test_that("best_global_ci_split_cpp matches R scoring with tied ranks", {
  predictors <- data.frame(
    z = c(1, 2, 3, 4, 5, 6, 7, 8),
    group = factor(c("a", "a", "b", "b", "c", "c", "d", "d"))
  )
  y <- cbind(
    rank = c(10, 20, 20, 40, 40, 60, 60, 80),
    outcome = c(0, 0, 1, 1, 4, 4, 5, 5)
  )
  wt <- c(1, 3, 2, 1, 4, 2, 5, 1)
  ctrl <- ci_tree_control(minsplit = 1, minbucket = 1, minprob = 0)

  current <- best_global_ci_split(
    x = predictors,
    y = y,
    wt = wt,
    ctrl = ctrl,
    ci_fun = ci_factory("CIg")
  )
  cpp <- best_global_ci_split_cpp(
    x = predictors,
    y = y,
    wt = wt,
    ctrl = ctrl,
    type = "CIg"
  )

  expect_equal(cpp$gain, current$gain, tolerance = 1e-12)
  expect_equal(cpp$varid, current$varid)
  expect_equal(cpp$left, current$left)
})
```

Run focused tests with:

```{r run-tests, eval=FALSE}
testthat::test_file("tests/testthat/test-ci_split_search_cpp.R")
```
