---
title: "Inequality-aware party trees"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r development, include=FALSE}
library(testthat)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!--
 Source of truth for the introductory pkgdown article.
 This file is intentionally written as a narrative vignette, not as a function
 reference page. Run the inflate chunk at the bottom after editing.
-->

```{r development-load}
pkgload::load_all(export_all = FALSE)
```

# Introduction

`ineqTrees` is designed for a specific kind of applied question: how is a
health outcome distributed across a socioeconomic ranking, and which subgroups
account for the remaining inequality?

The package keeps three pieces of information together:

- a socioeconomic rank, such as household wealth;
- a health outcome, such as under-five death;
- predictors that define interpretable subgroups.

The examples below use the `kenya` dataset shipped with the package. The data
are simulated from a DHS-style child survival setting, so they are useful for
demonstrating the workflow without requiring external survey files.

```{r setup}
library(ineqTrees)
library(data.table)

data(kenya, package = "ineqTrees")
setDT(kenya)
```

## 1. Prepare the analysis table

For a concentration-index tree, the response has two columns: the
socioeconomic ranking variable and the health outcome. Here, `wealth` is the
rank and `deadu5_num` is a binary indicator for under-five death.

```{r prepare-analysis-data}
analysis_vars <- c(
  "wealth",
  "deadu5_num",
  "rural",
  "ed",
  "reg",
  "unskilled",
  "sample_weight"
)

kenya_child <- kenya[
  complete.cases(kenya[, ..analysis_vars]),
  ..analysis_vars
]

kenya_child[
  ,
  .(
    children = .N,
    mortality = mean(deadu5_num),
    mean_wealth = mean(wealth)
  ),
  by = rural
]
```

For a fast vignette build we use a reproducible subset. In a real analysis you
would normally fit the model to the full prepared survey table.

```{r sample-analysis-data}
set.seed(20260512)
kenya_model <- kenya_child[
  sample.int(.N, min(1000L, .N))
]

predictors <- c("rural", "ed", "reg", "unskilled")
```

## 2. Measure socioeconomic inequality at the root

The concentration index summarizes whether the outcome is concentrated among
poorer or richer children after ordering observations by the ranking variable.
`ci_factory()` returns a scoring function. The same scoring function is used
inside the tree splitter.

```{r root-concentration-index}
ci <- ci_factory("CI")

root_ci <- ci(
  cbind(rank = kenya_model$wealth, outcome = kenya_model$deadu5_num),
  kenya_model$sample_weight
)

root_ci
```

The value is a root-node diagnostic: before any subgrouping, this is the
inequality score for the outcome in the analysis sample.

## 3. Fit a greedy concentration-index tree

`ci_tree()` grows a binary tree by looking for splits that reduce within-node
concentration-index impurity. The formula uses `cbind(wealth, deadu5_num)` so
the model knows both the ranking variable and the outcome.

```{r fit-ci-tree}
tree_fit <- ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  control = ci_tree_control(
    minsplit = 120,
    minbucket = 60,
    minprob = 0.05,
    maxdepth = 3
  )
)

tree_fit
```

The printed tree shows the model objective, the number of terminal nodes, and a
terminal-node summary. The most useful table for interpretation is often the
terminal summary itself.

```{r terminal-summary}
terminal_summary <- ci_tree_terminal_summary(tree_fit)

terminal_summary[
  order(-outcome_percent),
  .(node, n, depth, ci, outcome_percent, rule)
]
```

The `rule` column describes the subgroup path leading to each terminal node.
High mortality and high within-node concentration index point to subgroups
where the outcome remains both common and unequally distributed by wealth.

## 4. Plot the fitted tree

The plotting method uses `partykit` under the hood, but the package adds
compact split labels and terminal-node statistics that are easier to read for
large factor splits.

```{r plot-ci-tree, fig.width=12, fig.height=7, fig.alt="A concentration-index tree for under-five death in the Kenya example data."}
plot(
  tree_fit,
  data = kenya_model,
  var_labels = c(
    rural = "Residence",
    ed = "Mother education",
    reg = "Province",
    unskilled = "Mother occupation"
  ),
  plural_overrides = c(Province = "provinces"),
  terminal_stats = list(
    n = nrow,
    mortality = function(df) mean(df$deadu5_num),
    mean_wealth = function(df) mean(df$wealth)
  ),
  stat_labels = c(
    n = "n",
    mortality = "% death",
    mean_wealth = "mean wealth"
  ),
  stat_formatters = list(
    mortality = function(x) sprintf("%.1f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x)
  )
)
```

## 5. Use the tree as a prediction rule

The tree is fitted to explain inequality, but it can also be used as a simple
risk model by assigning each child the mean outcome in their terminal node.

```{r terminal-mean-predictions}
tree_risk <- predict_ci_tree_terminal_mean(
  fit = tree_fit,
  train_data = kenya_model,
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight
)

kenya_model[
  ,
  .(
    observed_mortality = mean(deadu5_num),
    mean_tree_risk = mean(tree_risk)
  )
]
```

The validation-gain score asks whether the fitted partition reduces
concentration-index impurity when observations are assigned to terminal nodes.

```{r tree-validation-gain}
ci_tree_validation_gain(
  fit = tree_fit,
  new_data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight
)
```

## 6. Tune tree controls

Tree depth and minimum node size control the trade-off between detail and
stability. `tune_ci_tree()` evaluates candidate controls by cross-validation.
The example below keeps the grid small so the vignette builds quickly.

```{r tune-tree}
tuning_grid <- ci_tree_control_grid(
  minsplit = c(120L, 180L),
  minbucket = c(60L),
  minprob = 0.05,
  maxdepth = 2:3
)

tree_tuning <- tune_ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = "CI",
  control_grid = tuning_grid,
  v = 3L,
  strata = "deadu5_num",
  seed = 20260512,
  refit = TRUE
)

tree_tuning$summary[
  order(-mean_score),
  .(minsplit, minbucket, maxdepth, mean_score, mean_terminal_nodes)
]
```

The selected tree is available as `tree_tuning$best_fit`, which can be printed,
summarized, or plotted in the same way as `tree_fit`.

```{r best-tuned-tree}
tree_tuning$best_params[
  ,
  .(minsplit, minbucket, maxdepth, mean_score, mean_terminal_nodes)
]
```

## 7. Fit a concentration-index forest

`ci_forest()` averages predictions from many greedy concentration-index trees.
The forest can be useful when the goal is a smoother risk estimate, while a
surrogate tree can still be fitted afterward for interpretation.

```{r fit-ci-forest}
set.seed(20260512)

forest_fit <- ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  ntree = 20L,
  mtry = 2L,
  control = ci_tree_control(
    minsplit = 120,
    minbucket = 60,
    minprob = 0.05,
    maxdepth = 3
  )
)

ci_forest_summary(forest_fit)
```

The fitted values are predicted risks. Their concentration index can be
compared with the concentration index of the observed outcome.

```{r forest-prediction-ci}
forest_risk <- predict(forest_fit, OOB = FALSE)

ci(
  cbind(rank = kenya_model$wealth, outcome = forest_risk),
  kenya_model$sample_weight
)
```

## 8. Decompose fitted inequality with SHAP values

When model explanations are available as SHAP values, `shap_conc_decomp()`
decomposes the concentration index of the fitted risk into feature-level
contributions. This step is model-agnostic: the SHAP values may come from an
`ineqTrees` forest, `ranger`, `xgboost`, or another prediction model.

The small example below uses an already-aligned SHAP-style contribution table.
In an applied workflow, this object would usually be produced by a SHAP
package.

```{r shap-decomposition-intro}
shap_example <- data.frame(
  residence = as.numeric(kenya_model$rural == "Rural") * 0.01,
  education = as.numeric(kenya_model$ed == "b no education") * 0.015,
  occupation = as.numeric(kenya_model$unskilled == "Yes") * 0.01
)

prediction <- mean(forest_risk) + rowSums(shap_example)

decomp <- shap_conc_decomp(
  shap = shap_example,
  rank = kenya_model$wealth,
  prediction = prediction,
  weights = kenya_model$sample_weight
)

decomp$contributions[
  order(-abs_contribution),
  .(feature, D_k_SHAP, pct_contribution)
]
```

This decomposition should be read as a model-based explanation of fitted
inequality, not as a causal attribution.

## 9. Where to go next

The introductory workflow covers the main objects:

- `ci_factory()` for concentration-index scoring;
- `ci_tree()` for interpretable inequality-aware subgrouping;
- `tune_ci_tree()` for choosing tree controls by validation gain;
- `ci_forest()` for ensemble risk estimation;
- `shap_conc_decomp()` for decomposing fitted inequality when SHAP values are
  available.

The reference pages document the lower-level split-search helpers and plotting
customization functions. Most analyses can start with the workflow shown here.

```{r development-inflate, eval=FALSE}
fusen::inflate(
  flat_file = "dev/flat_partytrees.Rmd",
  vignette_name = "Inequality-aware party trees",
  overwrite = "yes"
)
```
