---
title: "Inequality-aware party trees"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r development, include=FALSE}
library(testthat)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!--
 Source of truth for the introductory pkgdown article.
 This file is intentionally written as a narrative vignette, not as a function
 reference page. Run the inflate chunk at the bottom after editing.
-->

```{r development-load}
pkgload::load_all(export_all = FALSE)
```

# Introduction

`ineqTrees` is designed for a specific kind of applied question: how is a
health outcome distributed across a socioeconomic ranking, and which subgroups
account for the remaining inequality?

The package keeps three pieces of information together:

- a socioeconomic rank, such as household wealth;
- a health outcome, such as under-five death;
- predictors that define interpretable subgroups.

The examples below use the `kenya` dataset shipped with the package. The data
are simulated from a DHS-style child survival setting, so they are useful for
demonstrating the workflow without requiring external survey files.

```{r setup}
library(ineqTrees)
library(data.table)

data(kenya, package = "ineqTrees")
setDT(kenya)
```

## 1. Prepare the analysis table

For a concentration-index tree, the response has two columns: the
socioeconomic ranking variable and the health outcome. Here, `wealth` is the
rank and `deadu5_num` is a binary indicator for under-five death.

```{r prepare-analysis-data}
analysis_vars <- c(
  "wealth",
  "deadu5_num",
  "rural",
  "ed",
  "reg",
  "unskilled",
  "sample_weight"
)

kenya_child <- kenya[
  complete.cases(kenya[, ..analysis_vars]),
  ..analysis_vars
]

kenya_child[
  ,
  .(
    children = .N,
    mortality = mean(deadu5_num),
    mean_wealth = mean(wealth)
  ),
  by = rural
]
```

For a fast vignette build we use a reproducible subset. In a real analysis you
would normally fit the model to the full prepared survey table.

```{r sample-analysis-data}
set.seed(20260512)
kenya_model <- kenya_child[
  sample.int(.N, min(1000L, .N))
]

predictors <- c("rural", "ed", "reg", "unskilled")
criterion_types <- c("CI", "CIg", "CIc", "L")
```

## 2. Measure socioeconomic inequality at the root

The concentration index summarizes whether the outcome is concentrated among
poorer or richer children after ordering observations by the ranking variable.
`ci_root_impurity()` computes the unsplit/root objective with the same scoring
convention used by tree and forest tuning. Each criterion has its own root-node
benchmark.

```{r root-concentration-index}
whole_sample_ci <- data.table(type = criterion_types)[
  ,
  .(
    root_ci_index = ci_root_impurity(
      data = kenya_model,
      rank_name = "wealth",
      outcome_name = "deadu5_num",
      weights = kenya_model$sample_weight,
      type = type
    )
  ),
  by = type
]

ci <- ci_factory("CI")

whole_sample_ci
```

The `root_ci_index` is the baseline impurity before any subgrouping. The same
`type` argument appears in the tree and tuning functions, so each fitted model
should be compared with the root benchmark for the same concentration-index
criterion.

## 3. Fit trees with different concentration indices

`ci_tree()` can optimize several related concentration-index objectives. This
is important in applied work because the preferred index depends on the
outcome scale and on whether the ranking variable should be treated as a
fractional rank or as observed socioeconomic levels.

- `"CI"` is the standard concentration index.
- `"CIg"` is a generalized concentration index.
- `"CIc"` is a corrected concentration index for bounded outcomes.
- `"L"` uses the observed socioeconomic levels in the first response column.

You select the objective with the `type` argument. The fitted objects have the
same interface, so they can be printed, summarized, validated, or plotted in
the same way.

```{r fit-index-specific-trees}
tree_control <- ci_tree_control(
  minsplit = 120,
  minbucket = 60,
  minprob = 0.05,
  maxdepth = 3
)

tree_by_type <- setNames(
  lapply(criterion_types, function(ci_type) {
    ci_tree(
      formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
      data = kenya_model,
      rank_name = "wealth",
      outcome_name = "deadu5_num",
      weights = kenya_model$sample_weight,
      type = ci_type,
      control = tree_control
    )
  }),
  criterion_types
)

type_comparison <- rbindlist(lapply(names(tree_by_type), function(ci_type) {
  fit <- tree_by_type[[ci_type]]

  data.table(
    type = ci_type,
    terminal_nodes = length(partykit::nodeids(fit, terminal = TRUE)),
    validation_gain = ci_tree_validation_gain(
      fit = fit,
      new_data = kenya_model,
      rank_name = "wealth",
      outcome_name = "deadu5_num",
      weights = kenya_model$sample_weight,
      type = ci_type
    )
  )
}))

type_comparison[order(-validation_gain)]
```

In this quick comparison, larger validation gain means the terminal-node
partition removes more concentration-index impurity than the unsplit root in
the same sample. In a final analysis, use cross-validation to select among
indices and tree-size controls rather than choosing the largest in-sample
score.

For the rest of the introduction we continue with the standard concentration
index tree.

## 4. Fit a greedy concentration-index tree

`ci_tree()` grows a binary tree by looking for splits that reduce within-node
concentration-index impurity. The formula uses `cbind(wealth, deadu5_num)` so
the model knows both the ranking variable and the outcome.

```{r fit-ci-tree}
tree_fit <- ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = "CI",
  control = tree_control
)

tree_fit
```

The printed tree shows the model objective, the number of terminal nodes, and a
terminal-node summary. The most useful table for interpretation is often the
terminal summary itself.

```{r terminal-summary}
terminal_summary <- ci_tree_terminal_summary(tree_fit)

terminal_summary[
  order(-outcome_percent),
  .(node, n, depth, ci, outcome_percent, rule)
]
```

The `rule` column describes the subgroup path leading to each terminal node.
High mortality and high within-node concentration index point to subgroups
where the outcome remains both common and unequally distributed by wealth.

## 5. Plot the fitted tree

The plotting method uses `partykit` under the hood, but the package adds
compact split labels and terminal-node statistics that are easier to read for
large factor splits. The plain call is useful when you want a quick look at
the tree structure.

```{r plot-ci-tree-basic, fig.width=12, fig.height=7, fig.alt="A basic plot of a concentration-index tree for under-five death in the Kenya example data."}
plot(tree_fit)
```

For reports, the most useful improvement is usually to add readable variable
names and terminal-node summaries. The `terminal_stats`, `stat_labels`, and
`stat_formatters` arguments let the plot carry the sample size, outcome rate,
and mean rank in each terminal node instead of leaving those quantities for a
separate table.

```{r plot-ci-tree, fig.width=14, fig.height=10, fig.alt="A concentration-index tree for under-five death in the Kenya example data."}
plot(
  tree_fit,
  data = kenya_model,
  var_labels = c(
    rural = "Residence",
    ed = "Mother education",
    reg = "Province",
    unskilled = "Mother occupation"
  ),
  plural_overrides = c(Province = "provinces"),
  terminal_stats = list(
    n = nrow,
    mortality = function(df) mean(df$deadu5_num),
    mean_wealth = function(df) mean(df$wealth)
  ),
  stat_labels = c(
    n = "n",
    mortality = "% death",
    mean_wealth = "mean wealth"
  ),
  stat_formatters = list(
    mortality = function(x) sprintf("%.1f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x)
  )
)
```

With those arguments, the terminal panels become self-contained: a reader can
see not only where the split sends observations, but also how large each group
is, how common under-five death is in the group, and where the group sits on
the wealth ranking.

## 6. Use the tree as a prediction rule

The tree is fitted to explain inequality, but it can also be used as a simple
risk model by assigning each child the mean outcome in their terminal node.

```{r terminal-mean-predictions}
tree_risk <- predict_ci_tree_terminal_mean(
  fit = tree_fit,
  train_data = kenya_model,
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight
)

kenya_model[
  ,
  .(
    observed_mortality = mean(deadu5_num),
    mean_tree_risk = mean(tree_risk)
  )
]
```

The validation-gain score asks whether the fitted partition reduces
concentration-index impurity when observations are assigned to terminal nodes.

```{r tree-validation-gain}
ci_tree_validation_gain(
  fit = tree_fit,
  new_data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight
)
```

## 7. Select a tree by cross-validation

Tree depth, minimum node size, and the concentration-index objective all
control the trade-off between detail and stability. `tune_ci_tree()` is the
model-selection helper: it evaluates candidate controls by cross-validation,
scores each candidate on held-out folds, and can refit the selected tree on
the full analysis table.

The default metric is `"validation_gain"`, which selects the model that most
reduces concentration-index impurity on validation folds. Prediction-oriented
metrics are also available with `metric = "brier"`, `"log_loss"`, or
`"roc_auc"`. The relative gain metric scales validation gain by the fold's root
impurity, which is useful when comparing concentration-index criteria. The
example below keeps the grid small so the vignette builds quickly.

```{r tune-tree}
tuning_grid <- ci_tree_control_grid(
  minsplit = c(120L, 180L),
  minbucket = c(60L),
  minprob = 0.05,
  maxdepth = 2:3
)

tree_tuning <- tune_ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = criterion_types,
  control_grid = tuning_grid,
  v = 3L,
  strata = "deadu5_num",
  seed = 20260512,
  metrics = c("validation_gain", "relative_validation_gain"),
  refit = TRUE
)

tree_tuning_metrics <- ci_collect_metrics(
  tree_tuning,
  metric = c("validation_gain", "relative_validation_gain"),
  format = "wide"
)

tree_tuning_metrics[
  order(-mean_validation_gain),
  .(
    type,
    minsplit,
    minbucket,
    maxdepth,
    mean_validation_gain,
    mean_validation_relative_gain
  )
]
```

The tuning object stores fold-level metrics, grid-level summaries, selected
controls, and optional diagnostics. The collector helpers give a stable way to
inspect those pieces without relying on list internals.

```{r inspect-tree-tuning}
tree_selected <- ci_select_best(tree_tuning, metric = "validation_gain")

ci_fit_summary_table(
  tree_tuning,
  selected = tree_selected,
  metrics = c(
    "train_gain",
    "validation_gain",
    "train_relative_gain",
    "relative_validation_gain"
  )
)[
  ,
  .(
    type,
    minsplit,
    minbucket,
    maxdepth,
    mean_root_objective,
    mean_train_gain,
    mean_validation_gain,
    mean_validation_relative_gain,
    mean_percent_validation_gain
  )
]

tree_selected[
  ,
  .(type, minsplit, minbucket, maxdepth)
]
```

`control_ci_tune()` controls optional outputs. Here we score two metrics in a
single cross-validation run, save validation predictions, and extract a small
diagnostic from each fitted fold-level tree.

```{r tune-tree-control}
diagnostic_tuning <- tune_ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = "CI",
  control_grid = tuning_grid[1],
  v = 2L,
  strata = "deadu5_num",
  seed = 20260514,
  metrics = c("validation_gain", "brier"),
  control = control_ci_tune(
    save_pred = TRUE,
    extract = function(fit) {
      length(partykit::nodeids(fit, terminal = TRUE))
    }
  ),
  refit = FALSE
)

ci_collect_metrics(diagnostic_tuning, format = "wide")[
  ,
  .(
    type,
    mean_validation_gain,
    std_err_validation_gain,
    mean_brier,
    std_err_brier
  )
]

head(ci_collect_predictions(diagnostic_tuning)[
  ,
  .(grid_id, fold_id, row_id, outcome, .pred)
])

ci_collect_extracts(diagnostic_tuning)[
  ,
  .(
    grid_id,
    fold_id,
    type,
    terminal_nodes = unlist(.extracts)
  )
]
```

When you already use `rsample`, you can pass an rset with `resamples` instead
of asking `ineqTrees` to create folds from `v` and `strata`.

```{r tune-tree-rsample}
if (requireNamespace("rsample", quietly = TRUE)) {
  tree_folds <- rsample::vfold_cv(
    as.data.frame(kenya_model),
    v = 2L,
    strata = deadu5_num
  )

  rsample_tuning <- tune_ci_tree(
    formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
    data = kenya_model,
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    weights = kenya_model$sample_weight,
    type = "CI",
    control_grid = tuning_grid[1],
    resamples = tree_folds,
    seed = 20260515,
    refit = FALSE
  )

  ci_collect_metrics(rsample_tuning, format = "wide")[
    ,
    .(type, mean_validation_gain, n_validation_gain)
  ]
}
```

The top rows show the best candidates across the whole tuning run. Because the
four criteria are on different numerical scales, it is also useful to report
the best setting within each criterion and scale the validation gain by that
criterion's root impurity.

```{r best-tree-by-type}
tree_best_by_type <- ci_select_best(
  tree_tuning,
  metric = "relative_validation_gain"
)

tree_best_by_type_table <- ci_fit_summary_table(
  tree_tuning,
  selected = tree_best_by_type,
  metrics = c(
    "train_gain",
    "validation_gain",
    "train_relative_gain",
    "relative_validation_gain"
  )
)[
  ,
  .(
    type,
    minsplit,
    minbucket,
    minprob,
    maxdepth,
    mean_root_objective,
    mean_train_gain,
    mean_validation_gain,
    mean_validation_relative_gain,
    mean_percent_validation_gain
  )
]

tree_best_by_type_table[
  ,
  mean_percent_validation_gain := sprintf("%.1f%%", mean_percent_validation_gain)
]

tree_best_by_type_table
```

The selected tree is available as `tree_tuning$best_fit`, and the selected
index is available as `tree_tuning$best_type`. The best fit can be printed,
summarized, or plotted in the same way as `tree_fit`.

```{r best-tuned-tree}
tree_tuning$best_params[
  ,
  .(type, minsplit, minbucket, maxdepth)
]
```

## 8. Fit and tune a concentration-index forest

`ci_forest()` averages predictions from many greedy concentration-index trees.
The forest can be useful when the goal is a smoother risk estimate, while a
surrogate tree can still be fitted afterward for interpretation.

```{r fit-ci-forest}
set.seed(20260512)

forest_fit <- ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  ntree = 20L,
  mtry = 2L,
  control = ci_tree_control(
    minsplit = 120,
    minbucket = 60,
    minprob = 0.05,
    maxdepth = 3
  )
)

ci_forest_summary(forest_fit)
```

`ci_forest_surrogate()` fits an interpretable greedy tree to the fitted forest
predictions. Forest tuning still returns a surrogate for interpretation, while
the forest validation-gain metric itself is computed from the forest's stored
member trees.

```{r forest-surrogate}
forest_surrogate <- ci_forest_surrogate(
  forest_fit,
  data = kenya_model,
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  rank_name = "wealth",
  weights = kenya_model$sample_weight,
  prediction_name = "forest_risk",
  control = ci_tree_control(
    minsplit = 120,
    minbucket = 60,
    minprob = 0.05,
    maxdepth = 2
  )
)

ci_tree_terminal_summary(forest_surrogate$fit)[
  ,
  .(node, n, depth, outcome_percent, rule)
]
```

The same cross-validation idea applies to forests. `tune_ci_forest()` tunes
forest controls and computes validation gain by averaging the held-out
concentration-index gain across the forest's internal tree partitions.
Prediction metrics are computed from the averaged forest predictions, and the
selected refit still includes an interpretable surrogate tree. The grid below
uses `dials` to create a regular search over familiar tidymodels parameters,
then converts the result with `ci_dials_grid()`. For a larger applied analysis,
you could use
`levels = 3L` or add `dials::tree_depth()` and `dials::trees()` to the grid.

```{r tune-forest}
forest_dials_grid <- dials::grid_regular(
  dials::mtry(range = c(2L, length(predictors))),
  dials::min_n(range = c(60L, 120L)),
  levels = 2L
)

forest_grid <- ci_dials_grid(
  forest_dials_grid,
  minprob = 0.05,
  maxdepth = 2L,
  ntree = 8L
)

forest_tuning <- tune_ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = criterion_types,
  control_grid = forest_grid,
  v = 3L,
  strata = "deadu5_num",
  seed = 20260513,
  metrics = c("validation_gain", "relative_validation_gain", "brier"),
  control = control_ci_tune(save_pred = TRUE),
  refit = TRUE
)

forest_best_by_type <- ci_select_best(
  forest_tuning,
  metric = "relative_validation_gain"
)

forest_best_by_type_table <- ci_fit_summary_table(
  forest_tuning,
  selected = forest_best_by_type,
  metrics = c(
    "train_gain",
    "validation_gain",
    "train_relative_gain",
    "relative_validation_gain"
  )
)[
  ,
  .(
    type,
    ntree,
    mtry,
    minsplit,
    minbucket,
    minprob,
    maxdepth,
    mean_root_objective,
    mean_train_gain,
    mean_validation_gain,
    mean_validation_relative_gain,
    mean_percent_validation_gain
  )
]

forest_best_by_type_table[
  ,
  mean_percent_validation_gain := sprintf("%.1f%%", mean_percent_validation_gain)
]

forest_best_by_type_table
```

The same collector helpers work for forests. With `save_pred = TRUE`, the
validation predictions are available for quick diagnostic checks.

```{r inspect-forest-tuning}
ci_fit_summary_table(
  forest_tuning,
  selected = forest_best_by_type,
  metrics = c(
    "train_gain",
    "validation_gain",
    "train_relative_gain",
    "relative_validation_gain"
  )
)[
  ,
  .(
    type,
    ntree,
    mtry,
    minbucket,
    mean_validation_gain,
    mean_percent_validation_gain
  )
]

ci_collect_metrics(forest_tuning, metric = "brier", format = "wide")[
  order(mean_brier),
  .(type, ntree, mtry, mean_brier, std_err_brier)
]

head(ci_collect_predictions(forest_tuning)[
  ,
  .(grid_id, fold_id, type, row_id, outcome, .pred)
])
```

The percentage gain is calculated as
`100 * mean_validation_gain / abs(mean_root_objective)`. This shows how much of
the validation-fold root concentration-index impurity is recovered by the
selected model for each concentration-index criterion. Because the denominator
is criterion-specific, the percentage makes the gains easier to compare within
each CI definition.

The fitted values are predicted risks. Their concentration index can be
compared with the concentration index of the observed outcome.

```{r forest-prediction-ci}
forest_risk <- predict(forest_fit, OOB = FALSE)

ci(
  cbind(rank = kenya_model$wealth, outcome = forest_risk),
  kenya_model$sample_weight
)
```

## 9. Decompose fitted inequality with SHAP values

When model explanations are available as SHAP values, `shap_conc_decomp()`
decomposes the concentration index of the fitted risk into feature-level
contributions. This step is model-agnostic: the SHAP values may come from an
`ineqTrees` forest, `ranger`, `xgboost`, or another prediction model.

The small example below uses an already-aligned SHAP-style contribution table.
In an applied workflow, this object would usually be produced by a SHAP
package.

```{r shap-decomposition-intro}
shap_example <- data.frame(
  residence = as.numeric(kenya_model$rural == "Rural") * 0.01,
  education = as.numeric(kenya_model$ed == "b no education") * 0.015,
  occupation = as.numeric(kenya_model$unskilled == "Yes") * 0.01
)

prediction <- mean(forest_risk) + rowSums(shap_example)

decomp <- shap_conc_decomp(
  shap = shap_example,
  rank = kenya_model$wealth,
  prediction = prediction,
  weights = kenya_model$sample_weight
)

decomp$contributions[
  order(-abs_contribution),
  .(feature, D_k_SHAP, pct_contribution)
]
```

This decomposition should be read as a model-based explanation of fitted
inequality, not as a causal attribution.

## 10. Where to go next

The introductory workflow covers the main objects:

- `ci_factory()` for concentration-index scoring;
- `ci_tree()` for interpretable inequality-aware subgrouping with `type =
  "CI"`, `"CIg"`, `"CIc"`, or `"L"`;
- `tune_ci_tree()` for cross-validated model selection across tree controls
  and concentration-index objectives;
- `control_ci_tune()` and the `ci_collect_*()` helpers for saving predictions,
  extracting fold diagnostics, collecting notes, and inspecting tuning results;
- `ci_forest()` for ensemble risk estimation;
- `ci_forest_surrogate()` for fitting an interpretable tree to forest
  predictions;
- `ci_fit_summary_table()` for report-ready tuning summaries;
- `shap_conc_decomp()` for decomposing fitted inequality when SHAP values are
  available.

The reference pages document the lower-level split-search helpers and plotting
customization functions. Most analyses can start with the workflow shown here.

```{r development-inflate, eval=FALSE}
fusen::inflate(
  flat_file = "dev/flat_partytrees.Rmd",
  vignette_name = "Inequality-aware party trees",
  overwrite = "yes"
)
```
