---
title: "Shapley decomposition for inequality-aware trees"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r development, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This article shows how SHAP values can be connected to the concentration-index
decomposition in `ineqTrees`. The key point is model-agnostic:
`shap_conc_decomp()` only needs SHAP values, a socioeconomic rank, and model
predictions. Those predictions can come from an inequality-aware tree, an
inequality-aware forest fitted through tidymodels, or an ordinary prediction
model such as a `ranger` random forest.

The examples use the simulated Kenya child-survival data shipped with the
package. We fit three models:

- a package-native `ci_tree()` and `ci_forest()`;
- a tidymodels `rand_forest(engine = "ineqTrees")`;
- a standard `ranger` forest for predicting under-five mortality.

Then we approximate SHAP values with `fastshap::explain()`, decompose the
concentration index of the fitted risks, and plot the feature contributions.

```{r setup}
library(ineqTrees)
library(data.table)
library(parsnip)
library(workflows)
library(rsample)
library(dials)
library(tune)
library(yardstick)
library(hardhat)

register_ineqtrees_parsnip()

data(kenya, package = "ineqTrees")
setDT(kenya)
```

## 1. Prepare the analysis table

This chunk keeps the rank, outcome, predictors, and sample weight together.
The same table is used for the inequality-aware models and for the ordinary
prediction forest.

```{r prepare-analysis-data}
shap_predictors <- c("rural", "ed", "reg", "unskilled")

analysis_vars <- c(
  "wealth",
  "deadu5_num",
  shap_predictors,
  "sample_weight"
)

kenya_child <- kenya[
  complete.cases(kenya[, ..analysis_vars]),
  ..analysis_vars
]

set.seed(20260517)
kenya_model <- as.data.frame(
  kenya_child[sample.int(.N, min(800L, .N))]
)

kenya_model$case_wt <- hardhat::importance_weights(kenya_model$sample_weight)

model_x <- kenya_model[, shap_predictors, drop = FALSE]

eval_n <- min(250L, nrow(kenya_model))
eval_rows <- sort(sample.int(nrow(kenya_model), eval_n))
model_x_eval <- model_x[eval_rows, , drop = FALSE]
wealth_eval <- kenya_model$wealth[eval_rows]
weight_eval <- kenya_model$sample_weight[eval_rows]

head(kenya_model[, c("wealth", "deadu5_num", shap_predictors)])
```

## 2. Helper functions

This chunk defines prediction wrappers for `fastshap`. Each wrapper accepts a
model object and a predictor table, then returns one numeric risk for each row.

```{r helper-functions}
ci_tree_risk_predict <- function(object, newdata) {
  as.numeric(stats::predict(
    object,
    newdata = as.data.frame(newdata),
    type = "response"
  ))
}

ci_forest_risk_predict <- function(object, newdata) {
  as.numeric(stats::predict(
    object,
    newdata = as.data.frame(newdata),
    OOB = FALSE
  ))
}

ranger_risk_predict <- function(object, newdata) {
  prob <- predict(object, new_data = as.data.frame(newdata), type = "prob")
  as.numeric(prob$.pred_died)
}

plot_shap_contributions <- function(contributions, title) {

  ggplot2::ggplot(
    contributions,
    ggplot2::aes(
      x = stats::reorder(feature, pct_contribution),
      y = pct_contribution,
      fill = pct_contribution > 0
    )
  ) +
    ggplot2::geom_col(width = 0.7) +
    ggplot2::coord_flip() +
    ggplot2::scale_fill_manual(
      values = c("#2166ac", "#b2182b"),
      guide = "none"
    ) +
    ggplot2::labs(
      x = NULL,
      y = "Percentage contribution",
      title = title
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(panel.grid.minor = ggplot2::element_blank())
}
```

## 3. Fit package-native inequality-aware models

This chunk fits a single concentration-index tree and a small
concentration-index forest with the package-native API. These models optimize
concentration-index split gain, so they are explanatory subgrouping models
first and prediction models second.

```{r fit-native-models}
tree_fit <- ci_tree(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = "CI",
  control = ci_tree_control(
    minsplit = 120L,
    minbucket = 60L,
    minprob = 0.05,
    maxdepth = 3L
  )
)

set.seed(20260517)
forest_fit <- ci_forest(
  formula = cbind(wealth, deadu5_num) ~ rural + ed + reg + unskilled,
  data = kenya_model,
  rank_name = "wealth",
  outcome_name = "deadu5_num",
  weights = kenya_model$sample_weight,
  type = "CI",
  ntree = 20L,
  mtry = 2L,
  control = ci_tree_control(
    minsplit = 120L,
    minbucket = 60L,
    minprob = 0.05,
    maxdepth = 3L
  )
)

ci_forest_summary(forest_fit)
```

## 4. Decompose the inequality-aware tree

This chunk approximates SHAP values for the tree predictions and then
decomposes the concentration index of those fitted risks. If `fastshap` is not
installed, the chunk is skipped.

```{r tree-shap}
set.seed(20260517)

tree_pred_eval <- ci_tree_risk_predict(tree_fit, model_x_eval)

tree_shap <- fastshap::explain(
  object = tree_fit,
  X = model_x,
  pred_wrapper = ci_tree_risk_predict,
  newdata = model_x_eval,
  nsim = 32,
  adjust = TRUE
)

tree_decomp <- shap_conc_decomp(
  shap = tree_shap,
  rank = wealth_eval,
  prediction = tree_pred_eval,
  weights = weight_eval,
  type = "CI"
)

knitr::kable(
  as.data.frame(tree_decomp$diagnostics),
  digits = 3,
  caption = "CI tree SHAP decomposition diagnostics"
)
```

```{r tree-shap-table}
tree_contrib <- as.data.frame(tree_decomp$contributions)

knitr::kable(
  tree_contrib,
  digits = 3,
  caption = "CI tree SHAP-based concentration-index contributions"
)
```

```{r tree-shap-plot, fig.width=8, fig.height=5, fig.alt="A horizontal bar chart of CI tree SHAP-based concentration-index contributions."}
plot_shap_contributions(
  tree_contrib,
  "CI tree SHAP-based concentration-index decomposition"
)
```

## 5. Decompose the inequality-aware forest

The same decomposition works for a forest. The only difference is the
prediction wrapper: for `ci_forest()`, predictions are averaged across trees.

```{r forest-shap}
set.seed(20260518)

forest_pred_eval <- ci_forest_risk_predict(forest_fit, model_x_eval)

forest_shap <- fastshap::explain(
  object = forest_fit,
  X = model_x,
  pred_wrapper = ci_forest_risk_predict,
  newdata = model_x_eval,
  nsim = 32,
  adjust = TRUE
)

forest_decomp <- shap_conc_decomp(
  shap = forest_shap,
  rank = wealth_eval,
  prediction = forest_pred_eval,
  weights = weight_eval,
  type = "CI"
)

knitr::kable(
  as.data.frame(forest_decomp$diagnostics),
  digits = 3,
  caption = "CI forest SHAP decomposition diagnostics"
)
```

```{r forest-shap-table}
forest_contrib <- as.data.frame(forest_decomp$contributions)

knitr::kable(
  forest_contrib,
  digits = 3,
  caption = "CI forest SHAP-based concentration-index contributions"
)
```

```{r forest-shap-plot, fig.width=8, fig.height=5, fig.alt="A horizontal bar chart of CI forest SHAP-based concentration-index contributions."}
plot_shap_contributions(
  forest_contrib,
  "CI forest SHAP-based concentration-index decomposition"
)
```

## 6. Fit an inequality-aware forest with tidymodels

This chunk fits the same model family through parsnip and workflows. The
formula includes `wealth` so the parsnip bridge has the rank column available;
the bridge then removes `wealth` from the split predictors internally.

```{r fit-tidymodels-ineq-forest}
tidy_forest_spec <- rand_forest(
  trees = 20L,
  mtry = 2L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    maxdepth = 3L
  ) |>
  set_mode("regression")

tidy_forest_wf <- workflow() |>
  add_model(tidy_forest_spec) |>
  add_formula(deadu5_num ~ wealth + rural + ed + reg + unskilled) |>
  add_case_weights(case_wt)

set.seed(20260519)
tidy_forest_fit <- fit(tidy_forest_wf, data = kenya_model)
tidy_forest_engine <- extract_fit_parsnip(tidy_forest_fit)$fit

ci_forest_summary(tidy_forest_engine)
```

## 7. Decompose the tidymodels inequality-aware forest

The workflow formula includes `wealth` so the engine can learn inequality-aware
splits. For SHAP, we explain the fitted engine extracted from the workflow,
because the engine predicts from the split predictors without treating the
rank column as a determinant.

```{r tidy-forest-shap}
set.seed(20260519)

tidy_forest_pred_eval <- ci_forest_risk_predict(
  tidy_forest_engine,
  model_x_eval
)

tidy_forest_shap <- fastshap::explain(
  object = tidy_forest_engine,
  X = model_x,
  pred_wrapper = ci_forest_risk_predict,
  newdata = model_x_eval,
  nsim = 32,
  adjust = TRUE
)

tidy_forest_decomp <- shap_conc_decomp(
  shap = tidy_forest_shap,
  rank = wealth_eval,
  prediction = tidy_forest_pred_eval,
  weights = weight_eval,
  type = "CI"
)

knitr::kable(
  as.data.frame(tidy_forest_decomp$diagnostics),
  digits = 3,
  caption = "Tidymodels CI forest SHAP decomposition diagnostics"
)
```

```{r tidy-forest-shap-plot, fig.width=8, fig.height=5, fig.alt="A horizontal bar chart of tidymodels CI forest SHAP-based concentration-index contributions."}
tidy_forest_contrib <- as.data.frame(tidy_forest_decomp$contributions)

plot_shap_contributions(
  tidy_forest_contrib,
  "Tidymodels CI forest SHAP-based decomposition"
)
```

## 8. Fit a prediction-focused ranger forest

The decomposition does not require an inequality-aware model. This chunk fits
a standard `ranger` random forest whose target is prediction of under-five
mortality. The model is classification-focused; the decomposition is applied
afterward to the predicted mortality risk.

```{r ranger-model}
ranger_data <- kenya_model
ranger_data$deadu5 <- factor(
  ifelse(ranger_data$deadu5_num == 1, "died", "alive"),
  levels = c("died", "alive")
)

set.seed(20260520)
ranger_folds <- vfold_cv(ranger_data, v = 3L, strata = deadu5)

ranger_spec <- rand_forest(
  trees = 200L,
  mtry = tune(),
  min_n = tune()
) |>
  set_engine("ranger", probability = TRUE) |>
  set_mode("classification")

ranger_wf <- workflow() |>
  add_model(ranger_spec) |>
  add_formula(deadu5 ~ rural + ed + reg + unskilled)

ranger_grid <- grid_regular(
  mtry(range = c(1L, length(shap_predictors))),
  min_n(range = c(5L, 40L)),
  levels = 2L
)

set.seed(20260520)
ranger_tuned <- tune_grid(
  ranger_wf,
  resamples = ranger_folds,
  grid = ranger_grid,
  metrics = metric_set(roc_auc, accuracy)
)

best_ranger <- finalize_workflow(
  ranger_wf,
  select_best(ranger_tuned, metric = "roc_auc")
) |>
  fit(data = ranger_data)

collect_metrics(ranger_tuned)
```

## 9. Decompose the ranger prediction model

This chunk explains the ordinary prediction model, then decomposes inequality
in its predicted mortality risks. This is useful when the research question is:
which predictors account for socioeconomic inequality in model-predicted
mortality risk?

```{r ranger-shap}
set.seed(20260521)

ranger_x <- ranger_data[, shap_predictors, drop = FALSE]
ranger_x_eval <- ranger_x[eval_rows, , drop = FALSE]
ranger_pred_eval <- ranger_risk_predict(best_ranger, ranger_x_eval)

ranger_shap <- fastshap::explain(
  object = best_ranger,
  X = ranger_x,
  pred_wrapper = ranger_risk_predict,
  newdata = ranger_x_eval,
  nsim = 32,
  adjust = TRUE
)

ranger_decomp <- shap_conc_decomp(
  shap = ranger_shap,
  rank = ranger_data$wealth[eval_rows],
  prediction = ranger_pred_eval,
  weights = ranger_data$sample_weight[eval_rows],
  type = "CI"
)

knitr::kable(
  as.data.frame(ranger_decomp$diagnostics),
  digits = 3,
  caption = "Ranger SHAP decomposition diagnostics"
)
```

```{r ranger-shap-table}
ranger_contrib <- as.data.frame(ranger_decomp$contributions)

knitr::kable(
  ranger_contrib,
  digits = 3,
  caption = "Ranger SHAP-based concentration-index contributions"
)
```

```{r ranger-shap-plot, fig.width=8, fig.height=5, fig.alt="A horizontal bar chart of ranger SHAP-based concentration-index contributions."}
plot_shap_contributions(
  ranger_contrib,
  "Ranger SHAP-based concentration-index decomposition"
)
```

## 10. Interpretation

The same decomposition table can be read for all three model families. Positive
and negative contributions offset each other. The decomposition is model-based rather than
causal: it explains inequality in the fitted predictions, not the causal
effect of a determinant on mortality.

The important practical distinction is the model target:

- `ci_tree()` and `ci_forest()` build the inequality objective directly into
  the split search.
- `rand_forest(engine = "ineqTrees")` lets the same objective move through
  tidymodels workflows.
- `rand_forest(engine = "ranger")` is an ordinary mortality prediction model;
  `shap_conc_decomp()` then asks how its predicted risks are distributed across
  wealth.

```{r development-inflate, eval=FALSE}
fusen::inflate(
  flat_file = "dev/shap_article.Rmd",
  vignette_name = "Shapley decomposition",
  overwrite = "yes",
  open_vignette = FALSE,
  check = FALSE,
  document = FALSE,
  clean = "no"
)
```
