---
title: "Tidymodels workflows for inequality trees"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r development, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!--
 Source of truth for the tidymodels workflow article.
 This is a narrative vignette, not a function reference page.
-->

```{r development-load}
pkgload::load_all(export_all = FALSE)
```

# Introduction

This article shows how to fit `ineqTrees` models from the tidymodels
framework. The core idea is still the same as in the package-native API:
the model needs a socioeconomic rank, an outcome, and predictors that define
subgroups. The tidymodels bridge lets those pieces move through `parsnip`,
`workflows`, `rsample`, `dials`, and `tune`.

The examples use the simulated `kenya` data shipped with the package. We fit:

- a single concentration-index decision tree;
- a single concentration-index forest;
- plots for the fitted tree and an interpretable forest surrogate;
- tuned tree and forest workflows using `tune::tune_grid()`.

## 1. Load packages and data

This chunk loads the modeling packages used in the article. `ineqTrees`
registers its parsnip engines on load, and `register_ineqtrees_parsnip()` is
called explicitly so the code is clear when copied into a fresh session.

```{r setup}
library(ineqTrees)
library(parsnip)
library(workflows)
library(rsample)
library(dials)
library(tune)
library(yardstick)
library(hardhat)
library(data.table)

register_ineqtrees_parsnip()

data(kenya, package = "ineqTrees")
setDT(kenya)
```

## 2. Prepare a tidymodels analysis table

This chunk keeps the rank, outcome, predictors, and survey weight in one
complete analysis table. We then add `case_wt`, a hardhat importance-weight
column that workflows can pass into parsnip during fitting and tuning.

```{r prepare-analysis-data}
analysis_vars <- c(
  "wealth",
  "deadu5_num",
  "rural",
  "ed",
  "reg",
  "unskilled",
  "sample_weight"
)

kenya_model <- kenya[
  complete.cases(kenya[, ..analysis_vars]),
  ..analysis_vars
]

set.seed(20260513)
kenya_model <- as.data.frame(kenya_model[
  sample.int(.N, min(600L, .N))
])

kenya_model$case_wt <- hardhat::importance_weights(kenya_model$sample_weight)

predictor_formula <- deadu5_num ~ wealth + rural + ed + reg + unskilled

head(kenya_model[, c("wealth", "deadu5_num", "rural", "ed", "case_wt")])
```

The formula includes `wealth` on the right-hand side only so tidymodels keeps
the rank column available to the engine. The `ineqTrees` bridge removes
`wealth` from the split predictors internally and uses it as `rank_name`.

## 3. Fit a single CI decision tree

This chunk builds a parsnip decision-tree specification. The generic
tidymodels arguments map to `ineqTrees` controls: `tree_depth` maps to maximum
depth and `min_n` maps to the child-node size. Inequality-specific settings,
such as `rank_name`, `outcome_name`, and `type`, are supplied through
`set_engine()`.

```{r tree-spec}
tree_spec <- decision_tree(
  tree_depth = 3L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    min_gain = 0
  ) |>
  set_mode("regression")

tree_spec
```

This chunk fits the tree. The formula is the tidymodels-style single-outcome
formula; the bridge rebuilds the two-column `cbind(wealth, deadu5_num)`
response before calling `ci_tree()`.

```{r fit-tree}
tree_fit <- fit(
  tree_spec,
  predictor_formula,
  data = kenya_model,
  case_weights = kenya_model$case_wt
)

tree_fit$fit
```

This chunk creates fitted risks and terminal-node ids. Numeric predictions are
returned as a tibble with `.pred`; raw predictions return the terminal node
partition used for concentration-index gain.

```{r tree-predictions}
tree_pred <- predict(tree_fit, new_data = kenya_model)
tree_nodes <- predict(tree_fit, new_data = kenya_model, type = "raw")

tree_scores <- data.frame(
  truth = kenya_model$deadu5_num,
  pred = tree_pred$.pred,
  rank = kenya_model$wealth,
  node = tree_nodes,
  weight = kenya_model$sample_weight
)

ci_gain(
  tree_scores,
  truth = truth,
  estimate = pred,
  rank = rank,
  node = node,
  case_weights = weight,
  type = "CI"
)
```

## 4. Plot the fitted tree

This chunk plots the fitted `ci_tree` stored inside the parsnip fit. The plot
method comes from `ineqTrees` and `partykit`, so we can use the same reporting
options as package-native `ci_tree()` fits.

```{r plot-tree, fig.width=12, fig.height=7, fig.alt="A concentration-index decision tree fitted through parsnip."}
plot(
  tree_fit$fit,
  data = kenya_model,
  var_labels = c(
    rural = "Residence",
    ed = "Mother education",
    reg = "Province",
    unskilled = "Mother occupation"
  ),
  terminal_stats = list(
    n = nrow,
    mortality = function(df) mean(df$deadu5_num),
    mean_wealth = function(df) mean(df$wealth)
  ),
  stat_labels = c(
    n = "n",
    mortality = "% death",
    mean_wealth = "mean wealth"
  ),
  stat_formatters = list(
    mortality = function(x) sprintf("%.1f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x)
  )
)
```

## 5. Fit a single CI forest

This chunk builds a random-forest specification. The generic tidymodels
arguments map to forest controls: `trees` maps to `ntree`, `mtry` controls the
number of candidate split variables, and `min_n` maps to `minbucket`.

```{r forest-spec}
forest_spec <- rand_forest(
  trees = 20L,
  mtry = 2L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    maxdepth = 3L,
    perturb = list(replace = FALSE, fraction = 0.632)
  ) |>
  set_mode("regression")

forest_spec
```

This chunk fits the forest through parsnip and prints a compact summary of the
underlying `ci_forest` object.

```{r fit-forest}
set.seed(20260513)

forest_fit <- fit(
  forest_spec,
  predictor_formula,
  data = kenya_model,
  case_weights = kenya_model$case_wt
)

ci_forest_summary(forest_fit$fit)
```

This chunk predicts from the forest and measures inequality in the fitted
risks. This is not validation gain; it is the concentration index of the
forest's predicted outcome.

```{r forest-predictions}
forest_pred <- predict(forest_fit, new_data = kenya_model)

forest_scores <- data.frame(
  truth = kenya_model$deadu5_num,
  pred = forest_pred$.pred,
  rank = kenya_model$wealth,
  weight = kenya_model$sample_weight
)

ci_prediction_index(
  forest_scores,
  truth = truth,
  estimate = pred,
  rank = rank,
  case_weights = weight,
  type = "CI"
)
```

## 6. Plot the forest with a surrogate tree

A forest is an ensemble, so there is no single tree plot that fully describes
the model. This chunk fits a small surrogate CI tree to the forest predictions.
The surrogate is useful for communicating the main subgroup structure in the
forest risk surface.

```{r forest-surrogate}
surrogate_data <- kenya_model
surrogate_data$forest_risk <- forest_pred$.pred

surrogate_spec <- decision_tree(
  tree_depth = 3L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "forest_risk",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05
  ) |>
  set_mode("regression")

surrogate_fit <- fit(
  surrogate_spec,
  forest_risk ~ wealth + rural + ed + reg + unskilled,
  data = surrogate_data,
  case_weights = surrogate_data$case_wt
)

ci_tree_terminal_summary(surrogate_fit$fit)[
  ,
  c("node", "n", "ci", "outcome_mean", "rule")
]
```

This chunk plots the surrogate tree. Read it as a summary of the forest's
predicted risk pattern, not as the forest itself.

```{r plot-forest-surrogate, fig.width=12, fig.height=7, fig.alt="A surrogate concentration-index tree for forest predictions."}
plot(
  surrogate_fit$fit,
  data = surrogate_data,
  var_labels = c(
    rural = "Residence",
    ed = "Mother education",
    reg = "Province",
    unskilled = "Mother occupation"
  ),
  terminal_stats = list(
    n = nrow,
    mean_risk = function(df) mean(df$forest_risk),
    mean_wealth = function(df) mean(df$wealth)
  ),
  stat_labels = c(
    n = "n",
    mean_risk = "mean risk",
    mean_wealth = "mean wealth"
  ),
  stat_formatters = list(
    mean_risk = function(x) sprintf("%.1f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x)
  )
)
```

## 7. Tune a CI tree with `tune`

This chunk creates resamples. Because the outcome is binary, stratifying by
`deadu5_num` helps keep events distributed across folds.

```{r make-resamples}
set.seed(20260513)

folds <- vfold_cv(
  kenya_model,
  v = 3L,
  strata = deadu5_num
)

folds
```

This chunk creates a tunable tree specification. `tree_depth` and `min_n` are
marked with `tune()`, while the inequality-specific engine arguments remain
fixed for this search.

```{r tunable-tree-spec}
tree_tune_spec <- decision_tree(
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05
  ) |>
  set_mode("regression")
```

This chunk combines the model, formula, and case weights into a workflow.
Again, `wealth` is present in the formula to carry the rank through
tidymodels; the bridge removes it from the split predictors before fitting.

```{r tree-workflow}
tree_wf <- workflow() |>
  add_model(tree_tune_spec) |>
  add_formula(predictor_formula) |>
  add_case_weights(case_wt)

tree_wf
```

This chunk defines a small tuning grid. Larger applied analyses can expand
the ranges or use `grid_latin_hypercube()` for a broader search.

```{r tree-grid}
tree_grid <- grid_regular(
  tree_depth(range = c(2L, 4L)),
  min_n(range = c(40L, 90L)),
  levels = 2L
)

tree_grid
```

This chunk runs cross-validation with `tune_grid()`. The model is still an
inequality-aware tree; here the tuning metric is prediction error, which is
what `tune_grid()` can compute directly from `.pred` and the observed outcome.

```{r tune-tree}
set.seed(20260513)

tree_tuned <- tune_grid(
  tree_wf,
  resamples = folds,
  grid = tree_grid,
  metrics = metric_set(rmse, mae),
  control = control_grid(save_pred = TRUE)
)

collect_metrics(tree_tuned)
```

This chunk selects the best tree by RMSE and refits it on the full analysis
table. The resulting workflow can be predicted from like any other tidymodels
fit.

```{r finalize-tree}
best_tree <- select_best(tree_tuned, metric = "rmse")

final_tree_wf <- finalize_workflow(tree_wf, best_tree)
final_tree_fit <- fit(final_tree_wf, data = kenya_model)

best_tree
```

## 8. Tune a CI forest with `tune`

This chunk creates a tunable forest specification. Forest tuning usually
searches over the ensemble size, `mtry`, and the child-node size.

```{r tunable-forest-spec}
forest_tune_spec <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    maxdepth = 3L,
    perturb = list(replace = FALSE, fraction = 0.632)
  ) |>
  set_mode("regression")
```

This chunk wraps the forest specification in a workflow. The formula and case
weights are the same as for the tree workflow.

```{r forest-workflow}
forest_wf <- workflow() |>
  add_model(forest_tune_spec) |>
  add_formula(predictor_formula) |>
  add_case_weights(case_wt)
```

This chunk defines a deliberately small forest grid so the article builds
quickly. Increase `trees`, add more `levels`, or use randomized grids for a
final analysis.

```{r forest-grid}
forest_grid <- grid_regular(
  trees(range = c(8L, 16L)),
  mtry(range = c(1L, 3L)),
  min_n(range = c(40L, 90L)),
  levels = 2L
)

forest_grid
```

This chunk tunes the forest. The result has the standard `tune_results`
interface, so `collect_metrics()`, `collect_predictions()`, `select_best()`,
and `show_best()` work as expected.

```{r tune-forest}
set.seed(20260514)

forest_tuned <- tune_grid(
  forest_wf,
  resamples = folds,
  grid = forest_grid,
  metrics = metric_set(rmse, mae),
  control = control_grid(save_pred = TRUE, parallel_over = "resamples")
)

collect_metrics(forest_tuned)
```

This chunk selects and refits the best forest. The refitted object is a
workflow whose parsnip fit contains the underlying `ci_forest`.

```{r finalize-forest}
best_forest <- select_best(forest_tuned, metric = "rmse")

final_forest_wf <- finalize_workflow(forest_wf, best_forest)
final_forest_fit <- fit(final_forest_wf, data = kenya_model)

best_forest
```

This chunk extracts the fitted `ci_forest` from the final workflow and prints
the package-native forest summary.

```{r final-forest-summary}
final_forest_engine <- extract_fit_parsnip(final_forest_fit)$fit

ci_forest_summary(final_forest_engine)
```

## 9. Notes on tuning metrics

The `ineqTrees` engine optimizes concentration-index impurity while fitting.
The `tune` package then chooses hyperparameters from the predictions it
collects across resamples. Standard yardstick metrics such as RMSE and MAE
work directly in `tune_grid()`.

Exact CI validation gain needs a terminal-node partition in addition to
`.pred`. For a single fitted tree, obtain that partition with:

```{r raw-node-example}
head(predict(tree_fit, new_data = kenya_model, type = "raw"))
```

Then pass those node ids to `ci_gain()`, as shown earlier. For forests, raw
node predictions return one terminal-node column per tree, so a surrogate tree
is often easier to use when you want a single interpretable partition.

```{r development-inflate, eval=FALSE}
fusen::inflate(
  flat_file = "dev/tidmodels_ineqTrees.Rmd",
  vignette_name = "tidmodels_ineqTrees",
  overwrite = "yes",
  open_vignette = FALSE,
  check = FALSE,
  document = FALSE,
  clean = "no"
)
```
