---
title: "Tidymodels extension for inequality-aware trees"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r development, include=FALSE}
library(testthat)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


<!--
 Source of truth for the tidymodels extension article.
 This file is intentionally written as a narrative vignette, not as a function
 reference page. Run the inflate chunk at the bottom after editing.
-->

```{r development-load}
pkgload::load_all(export_all = FALSE)
```

# Introduction

`ineqTrees` is built around an applied question: how is a health outcome
distributed across a socioeconomic ranking, and which interpretable subgroups
account for the remaining inequality?

The package-native workflow in `ineqTrees` answers that question directly with
`ci_tree()`, `ci_forest()`, and the package's own tuning helpers. In some
projects, however, it is useful to fit the same models inside the tidymodels
framework so they can move through `parsnip`, `workflows`, `rsample`, `dials`,
and `tune`.

This article follows the same Kenya child-survival example used in the core
introductory vignette. It keeps the same substantive concepts:

- the root concentration index as the baseline inequality benchmark;
- concentration-index tree criteria such as `"CI"`, `"CIg"`, `"CIc"`, and
  `"L"`;
- interpretable tree plots and terminal-node summaries;
- forest risk estimates and model-based inequality summaries.

The difference is that the model specifications, resampling, and tuning are now
handled through tidymodels.

```{r setup}
library(ineqTrees)
library(parsnip)
library(workflows)
library(rsample)
library(dials)
library(tune)
library(yardstick)
library(hardhat)
library(data.table)

register_ineqtrees_parsnip()

data(kenya, package = "ineqTrees")
setDT(kenya)
```

## 1. Prepare the analysis table

The ingredients are the same as in the package-native workflow: a
socioeconomic rank, a health outcome, interpretable predictors, and case
weights.

```{r prepare-analysis-data}
analysis_vars <- c(
  "wealth",
  "deadu5_num",
  "rural",
  "ed",
  "reg",
  "unskilled",
  "sample_weight"
)

kenya_child <- kenya[
  complete.cases(kenya[, ..analysis_vars]),
  ..analysis_vars
]

kenya_child[
  ,
  .(
    children = .N,
    mortality = mean(deadu5_num),
    mean_wealth = mean(wealth)
  ),
  by = rural
]
```

For a fast article build we use a reproducible subset. The analysis object is
stored as a regular data frame because that is the most natural input type for
tidymodels workflows.

```{r sample-analysis-data}
set.seed(20260516)

kenya_model <- as.data.frame(
  kenya_child[sample.int(.N, min(800L, .N))]
)

kenya_model$case_wt <- hardhat::importance_weights(kenya_model$sample_weight)

predictors <- c("rural", "ed", "reg", "unskilled")
criterion_types <- c("CI", "CIg", "CIc", "L")
predictor_formula <- deadu5_num ~ wealth + rural + ed + reg + unskilled

plot_labels <- c(
  rural = "Residence",
  ed = "Mother education",
  reg = "Province",
  unskilled = "Mother occupation"
)

head(kenya_model[, c("wealth", "deadu5_num", "rural", "ed", "case_wt")])
```

The formula is written in the tidymodels style, with `deadu5_num` as the
outcome and `wealth` on the right-hand side. The `ineqTrees` parsnip bridge
uses `wealth` as `rank_name`, rebuilds the two-column concentration-index
response internally, and removes `wealth` from the split-search predictors.

## 2. Measure socioeconomic inequality at the root

The root concentration index is still the benchmark before any subgrouping. It
gives the level of whole-sample inequality that the tree or forest will try to
explain.

```{r root-concentration-index}
root_response <- cbind(
  rank = kenya_model$wealth,
  outcome = kenya_model$deadu5_num
)

root_weights <- kenya_model$sample_weight

whole_sample_ci <- data.table(criterion = criterion_types)[
  ,
  .(
    root_ci_index = ci_factory(criterion)(root_response, root_weights)
  ),
  by = criterion
]

whole_sample_ci
```

Each concentration-index criterion has its own scale, so fitted models should
always be compared with the root benchmark for the same `type`.

## 3. Fit trees with different concentration indices through tidymodels

The first question is the same one asked in the package-native vignette: what
happens if the tree is fitted under different concentration-index objectives?

Here the model specification is created with `decision_tree()` and
`set_engine("ineqTrees", ...)`, but the underlying fitted object is still a
`ci_tree`. That means the same validation and plotting methods remain
available after fitting.

```{r fit-index-specific-trees}
fit_tree_with_type <- function(ci_type) {
  spec <- decision_tree(
    tree_depth = 3L,
    min_n = 60L
  ) |>
    set_engine(
      "ineqTrees",
      rank_name = "wealth",
      outcome_name = "deadu5_num",
      type = ci_type,
      minsplit = 120L,
      minprob = 0.05,
      min_gain = 0
    ) |>
    set_mode("regression")

  fit(
    spec,
    predictor_formula,
    data = kenya_model,
    case_weights = kenya_model$case_wt
  )
}

tree_by_type <- setNames(
  lapply(criterion_types, fit_tree_with_type),
  criterion_types
)

type_comparison <- rbindlist(lapply(names(tree_by_type), function(ci_type) {
  fit_object <- tree_by_type[[ci_type]]$fit

  data.table(
    type = ci_type,
    terminal_nodes = length(partykit::nodeids(fit_object, terminal = TRUE)),
    validation_gain = ci_tree_validation_gain(
      fit = fit_object,
      new_data = kenya_model,
      rank_name = "wealth",
      outcome_name = "deadu5_num",
      weights = kenya_model$sample_weight,
      type = ci_type
    )
  )
}))

type_comparison[order(-validation_gain)]
```

This comparison is still on the inequality scale: larger validation gain means
the terminal-node partition removes more concentration-index impurity than the
unsplit root. In the next section we work with a single interpretable tree so
the mechanics of the tidymodels fit are easy to see.

## 4. Fit a single inequality-aware tree

`decision_tree()` supplies the generic tree controls. The engine-specific
arguments describe the inequality problem itself: which column is the rank,
which column is the outcome, and which concentration-index criterion should be
optimized.

```{r tree-spec}
tree_spec <- decision_tree(
  tree_depth = 3L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    min_gain = 0
  ) |>
  set_mode("regression")

tree_spec
```

```{r fit-ci-tree}
tree_fit <- fit(
  tree_spec,
  predictor_formula,
  data = kenya_model,
  case_weights = kenya_model$case_wt
)

tree_fit$fit
```

The fitted object inside the parsnip model fit is a `ci_tree`, so the usual
terminal-node summary remains available.

```{r terminal-summary}
terminal_summary <- ci_tree_terminal_summary(tree_fit$fit)

terminal_summary[
  order(-outcome_percent),
  .(node, n, depth, ci, outcome_percent, rule)
]
```

## 5. Plot the fitted tree

Because the underlying engine fit is still a `ci_tree`, the same plot method
works after a tidymodels fit.

The plain call is useful when you want a quick look at the tree structure.

```{r plot-ci-tree-basic, fig.width=12, fig.height=7, fig.alt="A basic concentration-index tree fitted through the tidymodels framework."}
plot(tree_fit$fit)
```

For reporting, the more useful plot is the enriched version with readable
labels and terminal-node summaries.

```{r plot-ci-tree, fig.width=12, fig.height=7, fig.alt="A concentration-index tree fitted through tidymodels for the Kenya under-five mortality example."}
plot(
  tree_fit$fit,
  data = kenya_model,
  var_labels = plot_labels,
  plural_overrides = c(Province = "provinces"),
  terminal_stats = list(
    n = nrow,
    mortality = function(df) mean(df$deadu5_num),
    mean_wealth = function(df) mean(df$wealth)
  ),
  stat_labels = c(
    n = "n",
    mortality = "% death",
    mean_wealth = "mean wealth"
  ),
  stat_formatters = list(
    mortality = function(x) sprintf("%.1f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x)
  )
)
```

The important point is that tidymodels does not take away the interpretability
machinery already built into `ineqTrees`; it just changes how the model is
specified and managed.

## 6. Use the fitted tree as a prediction rule

The tidymodels fit returns the same kinds of predictions you would expect from
other parsnip regression models. Numeric predictions come back as `.pred`, and
the raw tree partition can be recovered from the same fitted object.

```{r tree-predictions}
tree_pred <- predict(tree_fit, new_data = kenya_model)
tree_nodes <- as.integer(
  predict(tree_fit, new_data = kenya_model, type = "raw")
)

data.table::as.data.table(kenya_model)[
  ,
  .(
    observed_mortality = mean(deadu5_num),
    mean_tree_risk = mean(tree_pred$.pred)
  )
]
```

That raw node partition is what lets us reconnect the tidymodels fit to the
inequality interpretation. `ci_gain()` computes the concentration-index gain of
the fitted partition.

```{r tree-ci-gain}
tree_scores <- data.frame(
  truth = kenya_model$deadu5_num,
  pred = tree_pred$.pred,
  rank = kenya_model$wealth,
  node = tree_nodes,
  weight = kenya_model$sample_weight
)

ci_gain(
  tree_scores,
  truth = truth,
  estimate = pred,
  rank = rank,
  node = node,
  case_weights = weight,
  type = "CI"
)
```

## 7. Tune an inequality-aware tree with tidymodels

In the package-native workflow, `tune_ci_tree()` handles concentration-index
tree tuning directly. In tidymodels, the equivalent pieces are a tunable
specification, a workflow, a resampling object, and a tuning grid.

The resampling design is the same idea as before: keep the outcome reasonably
balanced across folds and score candidate models on held-out data.

```{r make-tree-resamples}
set.seed(20260516)

tree_folds <- vfold_cv(
  kenya_model,
  v = 3L,
  strata = deadu5_num
)

tree_folds
```

```{r tunable-tree-spec}
tree_tune_spec <- decision_tree(
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    min_gain = 0
  ) |>
  set_mode("regression")

tree_wf <- workflow() |>
  add_model(tree_tune_spec) |>
  add_formula(predictor_formula) |>
  add_case_weights(case_wt)

tree_grid <- grid_regular(
  tree_depth(range = c(2L, 4L)),
  min_n(range = c(40L, 90L)),
  levels = 2L
)

tree_grid
```

`tune_grid()` works on the prediction scale, so the simplest metrics inside the
tuning loop are standard regression scores such as RMSE and MAE. Exact
concentration-index gain can then be computed for the finalized tree, where the
terminal-node partition is available.

```{r tune-tree}
set.seed(20260516)

tree_tuned <- tune_grid(
  tree_wf,
  resamples = tree_folds,
  grid = tree_grid,
  metrics = metric_set(rmse, mae),
  control = control_grid(save_pred = TRUE)
)

collect_metrics(tree_tuned)
```

```{r inspect-tree-tuning}
show_best(tree_tuned, metric = "rmse", n = 4)

best_tree <- select_best(tree_tuned, metric = "rmse")
best_tree
```

After choosing the best settings, the workflow is finalized and refitted on the
full analysis table. The final workflow still contains a `ci_tree` engine fit,
so the fitted partition can again be scored with `ci_gain()`.

```{r finalize-tree}
final_tree_wf <- finalize_workflow(tree_wf, best_tree)
final_tree_fit <- fit(final_tree_wf, data = kenya_model)
final_tree_parsnip <- extract_fit_parsnip(final_tree_fit)

final_tree_scores <- data.frame(
  truth = kenya_model$deadu5_num,
  pred = predict(final_tree_fit, new_data = kenya_model)$.pred,
  rank = kenya_model$wealth,
  node = as.integer(
    predict(final_tree_parsnip, new_data = kenya_model, type = "raw")
  ),
  weight = kenya_model$sample_weight
)

ci_gain(
  final_tree_scores,
  truth = truth,
  estimate = pred,
  rank = rank,
  node = node,
  case_weights = weight,
  type = "CI"
)
```

## 8. Fit and visualize an inequality-aware forest

The forest is fitted through the same bridge, but now the generic tidymodels
arguments are `trees`, `mtry`, and `min_n`.

```{r forest-spec}
forest_spec <- rand_forest(
  trees = 20L,
  mtry = 2L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    maxdepth = 3L,
    perturb = list(replace = FALSE, fraction = 0.632)
  ) |>
  set_mode("regression")

forest_spec
```

```{r fit-forest}
set.seed(20260516)

forest_fit <- fit(
  forest_spec,
  predictor_formula,
  data = kenya_model,
  case_weights = kenya_model$case_wt
)

ci_forest_summary(forest_fit$fit)
```

The forest produces fitted risks. Because an ensemble has no single terminal
node assignment, we first keep the fitted risk table and then score a compact
surrogate tree with `ci_gain()`.

```{r forest-predictions}
forest_pred <- predict(forest_fit, new_data = kenya_model)

forest_scores <- data.frame(
  truth = kenya_model$deadu5_num,
  pred = forest_pred$.pred,
  rank = kenya_model$wealth,
  weight = kenya_model$sample_weight
)

head(forest_scores)
```

There is no single-tree object to plot for the forest itself, because the fit
is an ensemble. The practical way to visualize the forest is to fit a small
surrogate inequality-aware tree to the forest predictions and then plot that
tree.

```{r forest-surrogate}
surrogate_data <- kenya_model
surrogate_data$forest_risk <- forest_pred$.pred

surrogate_spec <- decision_tree(
  tree_depth = 3L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "forest_risk",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    min_gain = 0
  ) |>
  set_mode("regression")

surrogate_fit <- fit(
  surrogate_spec,
  forest_risk ~ wealth + rural + ed + reg + unskilled,
  data = surrogate_data,
  case_weights = surrogate_data$case_wt
)

ci_tree_terminal_summary(surrogate_fit$fit)[
  ,
  .(node, n, ci, outcome_mean, rule)
]
```

The surrogate tree gives the forest predictions an interpretable partition.
That partition can be scored with the same concentration-index gain metric used
for the single tree.

```{r forest-surrogate-ci-gain}
surrogate_nodes <- as.integer(predict(
  surrogate_fit,
  new_data = surrogate_data,
  type = "raw"
))

forest_gain_scores <- data.frame(
  truth = surrogate_data$deadu5_num,
  pred = surrogate_data$forest_risk,
  rank = surrogate_data$wealth,
  node = surrogate_nodes,
  weight = surrogate_data$sample_weight
)

ci_gain(
  forest_gain_scores,
  truth = truth,
  estimate = pred,
  rank = rank,
  node = node,
  case_weights = weight,
  type = "CI"
)
```

```{r plot-forest-surrogate, fig.width=12, fig.height=7, fig.alt="A surrogate concentration-index tree fitted to the forest predictions through tidymodels."}
plot(
  surrogate_fit$fit,
  data = surrogate_data,
  var_labels = plot_labels,
  plural_overrides = c(Province = "provinces"),
  terminal_stats = list(
    n = nrow,
    mean_risk = function(df) mean(df$forest_risk),
    mean_wealth = function(df) mean(df$wealth)
  ),
  stat_labels = c(
    n = "n",
    mean_risk = "mean risk",
    mean_wealth = "mean wealth"
  ),
  stat_formatters = list(
    mean_risk = function(x) sprintf("%.1f%%", 100 * x),
    mean_wealth = function(x) sprintf("%.2f", x)
  )
)
```

This plot should be read as a compact summary of the forest risk surface, not
as a literal picture of the full ensemble.

## 9. Tune an inequality-aware forest with tidymodels

Forest tuning follows the same workflow logic as tree tuning. The difference is
that the hyperparameter grid now spans the ensemble size, the number of
candidate split variables, and the child-node size.

```{r tunable-forest-spec}
forest_tune_spec <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "deadu5_num",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    maxdepth = 3L,
    perturb = list(replace = FALSE, fraction = 0.632)
  ) |>
  set_mode("regression")

forest_wf <- workflow() |>
  add_model(forest_tune_spec) |>
  add_formula(predictor_formula) |>
  add_case_weights(case_wt)

forest_grid <- grid_regular(
  trees(range = c(8L, 16L)),
  mtry(range = c(1L, 3L)),
  min_n(range = c(40L, 90L)),
  levels = 2L
)

forest_grid
```

```{r tune-forest}
set.seed(20260516)

forest_tuned <- tune_grid(
  forest_wf,
  resamples = tree_folds,
  grid = forest_grid,
  metrics = metric_set(rmse, mae),
  control = control_grid(save_pred = TRUE)
)

collect_metrics(forest_tuned)
```

```{r inspect-forest-tuning}
show_best(forest_tuned, metric = "rmse", n = 4)

best_forest <- select_best(forest_tuned, metric = "rmse")
best_forest
```

The finalized workflow again contains a regular `ci_forest` engine fit, so the
package-native summary methods still work after tuning.

```{r finalize-forest}
final_forest_wf <- finalize_workflow(forest_wf, best_forest)
final_forest_fit <- fit(final_forest_wf, data = kenya_model)
final_forest_engine <- extract_fit_parsnip(final_forest_fit)$fit

ci_forest_summary(final_forest_engine)
```

```{r tuned-forest-ci-gain}
final_forest_pred <- predict(final_forest_fit, new_data = kenya_model)$.pred

final_forest_data <- kenya_model
final_forest_data$forest_risk <- final_forest_pred

final_forest_surrogate_spec <- decision_tree(
  tree_depth = 3L,
  min_n = 60L
) |>
  set_engine(
    "ineqTrees",
    rank_name = "wealth",
    outcome_name = "forest_risk",
    type = "CI",
    minsplit = 120L,
    minprob = 0.05,
    min_gain = 0
  ) |>
  set_mode("regression")

final_forest_surrogate <- fit(
  final_forest_surrogate_spec,
  forest_risk ~ wealth + rural + ed + reg + unskilled,
  data = final_forest_data,
  case_weights = final_forest_data$case_wt
)

final_forest_nodes <- as.integer(predict(
  final_forest_surrogate,
  new_data = final_forest_data,
  type = "raw"
))

final_forest_gain_scores <- data.frame(
  truth = final_forest_data$deadu5_num,
  pred = final_forest_pred,
  rank = final_forest_data$wealth,
  node = final_forest_nodes,
  weight = final_forest_data$sample_weight
)

ci_gain(
  final_forest_gain_scores,
  truth = truth,
  estimate = pred,
  rank = rank,
  node = node,
  case_weights = weight,
  type = "CI"
)
```

## 10. Where to go next

The tidymodels workflow covers the same modeling ideas as the package-native
article, but through a different orchestration layer:

- `decision_tree()` and `rand_forest()` define generic model controls;
- `set_engine("ineqTrees", ...)` supplies the inequality-specific arguments;
- `fit()` returns parsnip model objects whose underlying engine fits are still
  `ci_tree` and `ci_forest` objects;
- `plot()` and `ci_tree_terminal_summary()` still work on the extracted tree
  fits;
- `workflow()`, `vfold_cv()`, `grid_regular()`, and `tune_grid()` provide the
  resampling and tuning machinery;
- `ci_gain()` reconnects the tidymodels fit to the concentration-index
  interpretation.

If you want the most direct concentration-index tuning workflow, the package
helpers such as `tune_ci_tree()` and `tune_ci_forest()` are still the most
natural choice. If you want model specifications, workflows, and resampling to
live inside tidymodels, the bridge shown here keeps that option open without
giving up the interpretation tools already built into `ineqTrees`.

```{r development-inflate, eval=FALSE}
fusen::inflate(
  flat_file = "dev/tidymodels_extension.Rmd",
  vignette_name = "Tidymodels extension",
  overwrite = "yes"
)
```
